### 2020-11-16 GenomeBiology/bmc_article.tex
initial import of old files

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{hyperref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Method}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
%   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{LM}\fnm{Leonardo} \snm{Morelli}}
\author[
   addressref={aff1,aff3},
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{VG}\fnm{Valentina} \snm{Giansanti}}
\author[
   addressref={aff1},
   corref={aff1},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
   email={cittaro.davide@hsr.it}
]{\inits{DC}\fnm{Davide} \snm{Cittaro}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Center for Omics Sciences, IRCCS San Raffaele Hospital}, % university, etc
  \street{Via Olgettina 58},                     %
  \postcode{20132}                                % post or zip code
  \city{Milan},                              % city
  \cny{Italy}                                    % country
}
\address[id=aff2]{%
  \orgname{{Università} Vita-Salute San Raffaele},
  \street{Via Olgettina 58},
  \postcode{20132}
  \city{Milan},
  \cny{Italy}
}
\address[id=aff3]{%
  \orgname{Dept. of Informatics, Systems and Communication, Univ. of Milan-Bicocca},
  \street{Viale Sarca 336},
  \postcode{20126}
  \city{Milan},
  \cny{Italy}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system. While properties of single cells is the primary endpoint of such analysis, these are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation. Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighborhood graphs by optimisation of modularity. In this paper we propose an alternative solution to this problem, based on nested Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.
%\parttitle{First part title} %if any
%Text for this section.

%\parttitle{Second part title} %if any
%Text for this section.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{Single cell}
\kwd{Cluster analysis}
\kwd{Stochastic Block Models}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Transcriptome analysis at single cell level by RNA sequencing (scRNA-seq) is a technology growing in popularity and applications \cite{svensson_2018}. It has been applied to study the biology of complex tissues \cite{guo_2018, ventotormo_2018}, tumor dynamics \cite{rozenblattrosen_2020, tirosh_2016, patel_2014, neftel_2019}, development \cite{rosenberg_2018, wagner_2018} and to describe whole organisms \cite{plass_2018, regev_2017}.

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighborhood graph (\emph{e.g.} by \emph{k} Nearest Neighbors, \emph{k}NN) which is then partitioned into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee that small communities in large networks are well defined. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy and PhenoGraph \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006} . In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Additional solutions to cell group identification from neighborhood graphs have been proposed, introducing resampling techniques \cite{baran_2019} or clique analysis \cite{xu_2015}. Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organized into communities \cite{holland_1983}. In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed neighborhood graph. According to the microcanonical formulation \cite{peixoto_2017}, the parameters are node partitions into groups and the matrix of edge counts between groups themselves. Under this model, nodes belonging to the same group have the same probability to be connected with other nodes. It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs. A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \emph{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}). There, a generative model of network $\boldsymbol A$ has a probability $P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)$ where \textbf{$\boldsymbol\theta$} is the set of parameters and \textbf{\emph{$\boldsymbol b$}} is the set of partitions. The likelihood of the network being generated by a given partition can be measured by the posterior probability

\begin{equation}
P(\boldsymbol b | \boldsymbol A) = \frac{P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)P(\boldsymbol\theta, \boldsymbol b)}{P(\boldsymbol A)}
\end{equation}

and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length

\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}

so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo (MCMC) approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

In this work we propose nSBM for the analysis of single cell data, in particular scRNA-seq data. Our approach identifies cell groups in a statistical robust way and, moreover, is able to determine the likelihood of the grouping, thus allowing model selection. In addition, our approach measures the confidence of assignment to groups; we show that this information may be exploited to perfect the notion of cell groups and the identification of markers.

Lastly, we developed \emph{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \emph{scanpy}, to facilitate the adoption of nested stochastic block models in single-cell analysis.

\section*{Results}

\subsection*{Overview of \emph{schist}}

\emph{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \emph{scanpy}. The most prominent function is \emph{schist.inference.nested\_model()} which takes a \emph{AnnData} object as input and fits a nested stochastic block model on the \emph{k}NN graph built with \emph{scanpy} functions (\emph{e.g.\ scanpy.tools.neighbors()}). When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \emph{schist} annotates cells in the data object with all the groups found at each level of a hierarchy. As there could be more model fits with similar entropy, \emph{schist} could explore the space of solutions with a Markov Chain Monte Carlo algorithm, to perform model averaging; this step is performed until it converges, that is the difference in model entropy in \emph{n} continuous iterations remains under a specified threshold. Sampling from the posterior distribution can be used to study the distribution of the number of groups, at each level. This information (group marginals) could be studied to identify the most probable number of cell groups.

Once \emph{schist} has fitted a model, it evaluate the difference in entropy given by assigning every cell to every possible group. This step generates the matrix of \emph{cell affinity}, that is the probability for a cell to belong to a specific group. A cell affinity matrix is generated and returned for every hierarchy level, here including level 0. As we show below, cell affinities could be exploited as covariates in testing marker genes and, more in general, to define the stability of any cell group.

\subsection*{nSBM correctly identifies cell populations}
We tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on the mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighborhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbors in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbors. In general, higher number of components and neighbors has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbors. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbors are chosen, the graph is sparser and the model is fit from less edges (Fig. S1). Running MCMC algorithm recovers the performances of the majority of the configurations (Fig. S2).

Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig. \ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$). In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3. Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them. It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig. \ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.

These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter. These data also hint at the possibility to identify rare cell types in larger populations.

\subsection*{nSBM hierarchy contains biological information}
The hierarchical model of cell groups implies that a relationship exists between groups. We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation. To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig. S3A). After applying nSBM, we identify 27 groups at level 3 of the hierarchy (Fig. S3B), compared to the 24 using Leiden method at default resolution (Fig. S4). We found that the hierarchy proposed by our model is consistent with the developmental model (Fig. \ref{Figure2}). Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (Fig. S4); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.

In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.

\subsection*{Cell affinities can be used to evaluate cluster purity}
The computational framework underlying \emph{schist} calculates the model entropy, that is the amount of information required to describe a block configuration. Given that minimisation of such quantity can be used to perform model selection, it can be also used to evaluate the impact of modifying the assignment of a cell to a cluster. Once a model is minimised, \emph{schist} performs an exhaustive exploration of all model entropies resulting from moving all cells into all possible clusters. The differences in entropies could be interpreted as affinities of cells to given clusters. Such affinities are, in fact, probability values and could be used to evaluate the internal consistency of a given cell cluster.

To this end we calculate the entropy of the group-wise distribution of cell affinities, which is maximal when all cells have affinity equal to 1 for a given group. We tested this idea on four datasets recently published to benchmark single cell technologies in the Human Cell Atlas project \cite{mereu_2020}; in particular, we chose two technologies resulting in high quality data: Quartz-seq2 \cite{sasagawa_2018} and Chromium 10x v3 \cite{zheng_2017}, and two technologies resulting in more noisy data: MARS-seq \cite{jaitin_2014} and iCell8 \cite{goldstein_2017} (Fig. \ref{Figure3}).

Cluster consistency is not a measure of the data quality, in fact we identify low consistency groups in all datasets. High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper. Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells (Fig. S5).

Similarly, we can use cell affinities to derive a stability parameter, a measure of the tendency for a cell to be stably associated to given clusters at all levels of the hierarchy. To this end, we first calculate the cell-wise entropy $H_{i,h}$ of cell affinity at each hierarchy level \emph{h}, then we define the stability as $S_i = 1 - \max(H_i)$. While we conceived this measure to identify and exclude cells with dubious assignment, we found that it may be more useful to assess the general data quality: the fraction of cells having $S>0.95$ was 0.783, 0.795, 0.831 and 0.855 for the iCELL8, MARS-seq, Chromium 10x and Quartz-seq2 technology respectively, in line with the evaluation on increasing performances of those platforms in \cite{mereu_2020}.

\section*{Conclusions}

Identification of cells sharing similar properties in single cell experiments is of paramount importance. A large number of approaches have been described, although the standardisation of analysis pipelines converged to methods that are based on modularity optimisation. We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques. The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution. In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble. Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity. Lastly, a statistical way to evaluate models is made available, allowing for reliable model selection. This last capability has the obvious advantage that the choice of parameters, hence the definition of cell clusters, could be conditioned to an evaluation metric which is robust and easy to understand (\emph{i.e.} the model entropy).

The major drawback of adopting this strategy is the substantial increase of runtimes. According to the developers of the underlying libraries, runtimes are proportional to the number of edges in the neighborhood graph and while it supports CPU-level parallelisation, a model minimisation is hundreds times slower than  the extremely fast Leiden approach. Runtimes are further inflated if MCMC equilibration is performed. We are well aware that this will limit the adoption of any strategy based on nSBM, but we believe that the quality of the results greatly justifies the additional resources.

\section*{Materials and Methods}

A detailed view of the parameters used for the analysis presented in this manuscript is available as jupyter notebook at \href{https://github.com/dawe/schist-notebooks/tree/master/schist_paper}{https:/\slash github.com\slash dawe\slash schist-notebooks\slash tree\slash master\slash schist\_paper}.

\subsection*{Analysis of cell mixtures}

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Data were analysed using scanpy v1.4.6 \cite{wolf_2018}. Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. nSBM was initialised three times.

\subsection*{Analysis of hematopoietic differentiation}

Data were retrieved using scanpy's built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbors and diffmap as embedding. nSBM was completed with 3 initialisations. Gene signatures were calculated using the following gene lists
\begin{itemize}
\item Erythroids: Gata1, Klf1, Epor, Gypa, Hba-a2, Hba-a1, Spi1
\item Neutrophils, Elane, Cebpe, Ctsg, Mpo, Gfi1
\item Monocytes, Irf8, Csf1r, Ctsg, Mpo
\end{itemize}

\subsection*{Analysis of cluster consistency}

Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8). Data were processed according to the methods in the original paper \cite{mereu_2020}. Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded. Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses. Genes that were expressed in less than five cells were removed. Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5. Neighborhood graph was built using 30 principal components and 30 neighbors. nSBM was completed with 3 initialisations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
LM performed the analysis, contributed to the code and wrote the manuscript. VG supervised the analysis and wrote the manuscript. DC conceived the project, performed the analysis, contributed to the code and wrote the manuscript.

\section*{Acknowledgements}
The authors want to thank Tiago Peixoto and Giovanni Petri for ther inputs and suggestions.  The authors also want to thank Giovanni Tonon, Paolo Provero and all people at COSR for comments and support.
This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
\begin{figure}[h!]
\centering
%\includegraphics[keepaspectratio]{Figure_1.png}
\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text.}\label{Figure1}
\end{figure}

\begin{figure}[h!]
\centering
%\includegraphics[keepaspectratio]{Figure_2.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
\end{figure}

\begin{figure}[h!]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.7\textheight]{Figure_3.png}
\caption[]{Analysis of cell cluster consistency. Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform. Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1. The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster. Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
%\section*{Tables}
%\begin{table}[h!]
%\caption{Sample table title. This is where the description of the table should go.}
%      \begin{tabular}{cccc}
%        \hline
%           & B1  &B2   & B3\\ \hline
%        A1 & 0.1 & 0.2 & 0.3\\
%        A2 & ... & ..  & .\\
%        A3 & ..  & .   & .\\ \hline
%      \end{tabular}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Supplementary figures}
  The file supplementary.pdf includes supplementary figures from Figure S1 to Figure S5

\end{backmatter}
\end{document}

### 2020-11-16 GenomeBiology/supplementary.tex
initial import of old files

\documentclass[11pt, titlepage, twoside]{article}
\usepackage[top=20mm, bottom=20mm, left=20mm, right=20mm]{geometry}
\geometry{a4paper}
\geometry{portrait}
\usepackage[utf8]{inputenc}	% Comment out for xelatex; must use with pdflatex
\usepackage[T1]{fontenc}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{fontspec}
\usepackage{epstopdf}
\usepackage{float}
\usepackage[hang]{footmisc}
\usepackage{endnotes}
\usepackage{listingsutf8}
%\restylefloat{table}
\usepackage{xcolor}
\usepackage{colortbl}
%\usepackage{booktabs}
\usepackage{tabu}
\usepackage{caption}
\usepackage{textcomp}
\usepackage{textgreek}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\usepackage{authblk}

\makeatletter
\def\blx@maxline{77}
\makeatother

\captionsetup[figure]{labelfont=sc}
\captionsetup[table]{labelfont=sc}
\captionsetup[tabu]{labelfont=sc}

\DeclareCaptionType{MPEquation}[][List of equations]
\captionsetup[MPEquation]{labelformat=empty}

\DeclareCaptionType{MPListing}[][List of listings]
\captionsetup[MPListing]{labelformat=empty}

\usepackage{fancyhdr}
\setlength{\headheight}{16pt}
\pagestyle{fancyplain}
\fancyhf{}
%\fancypagestyle{plain}{}
\cfoot[\thepage]{\thepage}

\usepackage{stringenc}
\usepackage{pdfescape}

% Helper for debugging pdflatex errors induced by invalid UTF8 characters:
% output the actual character code causing the issue.

\makeatletter
\renewcommand*{\UTFviii@defined}[1]{%
\ifx#1\relax
\begingroup
% Remove prefix "\u8:"
\def\x##1:{}%
% Extract Unicode char from command name
% (utf8.def does not support surrogates)
\edef\x{\expandafter\x\string#1}%
\StringEncodingConvert\x\x{utf8}{utf16be}% convert to UTF-16BE
% Hexadecimal representation
\EdefEscapeHex\x\x
% Enhanced error message
\PackageError{inputenc}{Unicode\space char\space \string#1\space
(U+\x)\MessageBreak
not\space set\space up\space
for\space use\space with\space LaTeX}\@eha
\endgroup
\else\expandafter
#1%
\fi
}
\makeatother

% Make multi-paragraph footnotes align nicely on the left
\setlength\footnotemargin{10pt}

% Do not display an automatic Notes heading for endnotes, as they are contained
% by a Manuscripts section with a heading that has possibly been edited by the user
\def\enoteheading{}

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{document}
\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle
\section*{Supplementary figures}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors. Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis). Both the parameters influence the sparseness of the graph.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
\caption[]{Analysis of hematopoietic differentiation. (A) Low dimensional embedding of single cells colored by original cell type and pseudotime. (B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S4.png}
\caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1. Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution $r=0.3$ splits in groups 2 and 0 at resolution $r=0.2$.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency. Dot size is proportional to the number of cells remaining after filtering.}
\end{figure}

\end{document}

### 2020-11-16 PlosCompBio/plos_latex_template.tex
initial import of old files

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Nested stochastic block models applied to the analysis of single cell data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Leonardo Morelli\textsuperscript{1,2},
Valentina Giansanti\textsuperscript{1, 3},
Davide Cittaro\textsuperscript{1*},
\\
\bigskip
\textbf{1} Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy
\\
\textbf{2} Universit\`{a} Vita-Salute San Raffaele, Milan, Italy
\\
\textbf{3} Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* cittaro.davide@hsr.it

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system. While properties of single cells is the primary endpoint of such analysis, these are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation. Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity. In this paper we propose an alternative solution to this problem, based on nested Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Identification of cell types is a key step of many single cell experiments. Many of the approaches to achieve this goal converge on the analysis of the graph representing cell-wise similarity by optimisation of modularity. While fast, this method does not provide a robust estimation of the number of groups, and it leaves large room of arbitrariness in the choice of appropriate resolution. To overcome these limitations, we propose a strategy based on nested stochastic block models. We show our method is accurate and provides a rich description of cell groups and their relationships.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Transcriptome analysis at single cell level by RNA sequencing (scRNA-seq) is a technology growing in popularity and applications \cite{svensson_2018}. It has been applied to study the biology of complex tissues \cite{guo_2018, ventotormo_2018}, tumor dynamics \cite{rozenblattrosen_2020, tirosh_2016, patel_2014, neftel_2019}, development \cite{rosenberg_2018, wagner_2018} and to describe whole organisms \cite{plass_2018, regev_2017}.

A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, groups of cells sharing similar properties. Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}. As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards. Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN) which is then partitioned into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity. While fast, this method does not guarantee that small communities in large networks are well defined. To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy and PhenoGraph \cite{levine_2015}. In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006} . In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}. Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter. In addition, the larger the dataset, the harder is to identify small cell groups, as a consequence of the well-known resolution limit \cite{fortunato_2007}. Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model. Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019} or clique analysis \cite{xu_2015}. Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.

An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organized into communities \cite{holland_1983}. In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed neighbourhood graph. According to the microcanonical formulation \cite{peixoto_2017}, the parameters are node partitions into groups and the matrix of edge counts between groups themselves. Under this model, nodes belonging to the same group have the same probability to be connected with other nodes. It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs. A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \emph{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}). There, a generative model of network $\boldsymbol A$ has a probability $P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)$ where \textbf{$\boldsymbol\theta$} is the set of parameters and \textbf{\emph{$\boldsymbol b$}} is the set of partitions. The likelihood of the network being generated by a given partition can be measured by the posterior probability

%\begin{MPEquation}[!ht]
\begin{equation}
P(\boldsymbol b | \boldsymbol A) = \frac{P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)P(\boldsymbol\theta, \boldsymbol b)}{P(\boldsymbol A)}
\end{equation}
\label{MPEquationElement:820CB015-E51E-4C5A-BD9C-70E667B8F73B}
%\end{MPEquation}
and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length

%\begin{MPEquation}[!ht]
\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}
\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
%\end{MPEquation}
so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

In this work we propose nSBM for the analysis of single cell data, in particular scRNA-seq data. This approach identifies cell groups in a statistical robust way and, moreover, it is able to determine the likelihood of the grouping, thus allowing model selection. In addition, it is possible to measure the confidence of assignment to groups. We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.

We developed \emph{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \emph{scanpy}, to facilitate the adoption of nested stochastic block models in single-cell analysis.

\section*{Materials and methods}
%\subsection*{Analysis of Random data}
%Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function. The random \emph{k}NN graph was obtained shuffling the node labels of each edge. nSBM was performed with 3 initialisations using merge-split MCMC approach. UMAP embedding was recomputed after randomisation using the shuffled graph.

\subsection*{Analysis of cell mixtures}

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Data were analysed using scanpy v1.4.6 \cite{wolf_2018}. Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. nSBM was initialised three times. Analysis was performed at level 2 of the nSBM hierarchy.
Random Poisson noise was generated at given $\lambda$ (range: 0 - 1000) and added to the initial count matrix. Analysis was performed at level 2 of the nSBM hierarchy.

\subsection*{Analysis of hematopoietic differentiation}

Data were retrieved using scanpy's built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbours and diffmap as embedding. Gene signatures were calculated using the following gene lists
\begin{itemize}
\item Erythroids: Gata1, Klf1, Epor, Gypa, Hba-a2, Hba-a1, Spi1
\item Neutrophils, Elane, Cebpe, Ctsg, Mpo, Gfi1
\item Monocytes, Irf8, Csf1r, Ctsg, Mpo
\end{itemize}

nSBM was completed with 3 initialisations

\subsection*{Analysis of cluster consistency}

Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8). Data were processed according to the methods in the original paper \cite{mereu_2020}. Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded. Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses. Genes that were expressed in less than five cells were removed. Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5. Neighbourhood graph was built using 30 principal components and 20 neighbours. nSBM was completed with 3 initialisations.

% Results and Discussion can be combined.
\subsection*{Overview of \emph{schist}}

\emph{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \emph{scanpy}. The most prominent function is \emph{schist.inference.nested\_model()} which takes a \emph{AnnData} object as input and fits a nested stochastic block model on the \emph{k}NN graph built with \emph{scanpy} functions (\emph{e.g.\ scanpy.tools.neighbors()}). When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph. \emph{schist} annotates cells in the data object with all the groups found at each level of a hierarchy. As there could be more model fits with similar entropy, \emph{schist} could explore the space of solutions with a Markov Chain Monte Carlo (MCMC) algorithm, to perform model averaging; this step is performed until it converges, that is the difference in model entropy in \emph{n} continuous iterations remains under a specified threshold. Sampling from the posterior distribution can be used to study the distribution of the number of groups, at each level. This information (group marginals) could be studied to identify the most probable number of cell groups.
Once \emph{schist} has fitted a model, it evaluates the difference in entropy given by assigning every cell to every possible group. This step generates the matrix of \emph{cell affinity}, that is the probability for a cell to belong to a specific group. A cell affinity matrix is generated and returned for every hierarchy level. We show that this information can be useful to evaluate cluster consistence.

%As the core object of \emph{graph-tool }used by \emph{schist}, the \emph{NestedBlockState}, cannot be serialised by \emph{scanpy}'s IO infrastructure, we implemented simple read\slash write functions to dump \emph{AnnData} is h5ad format, together with the \emph{NestedBlockState} as a separate file using python's \emph{pickle} library.

%\subsection*{nSBM does not return artificial cell groupings}
%One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0431$, $p=0.993$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.

%This experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.

%\begin{figure}[!h]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
%\caption[]{nSBM results for randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches ($ARI=0.804$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
%\end{figure}

\subsection*{nSBM correctly identifies cell populations}
To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours. In general, higher number of components and neighbours has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (\nameref{S1_Fig}). Running MCMC algorithm recovers the performances of the majority of the configurations (\nameref{S2_Fig}).

Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig. \ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$). In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3. Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them. It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig. \ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.

It is worth noting that since nSBM reflects the amount of information contained in the data, it does not return clusters when data are noisy. To demonstrate this property, we added increasing levels of random Poisson noise to the count matrix (Fig.\ref{Figure_Random}). The number of groups reported by nSBM decreases with increasing amount of noise and it is 1 (\emph{i.e.} no clusters) at $\lambda=200$, whereas a standard approach overfits data.

These observations show that nSBM is suitable for accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter and with the possibility to identify rare cell types in larger populations. Moreover, nSBM avoids excessive clustering of data in presence of noisy measurements.

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
\end{figure}

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Tian_Random.png}
\caption[]{Identification of cell groups at different level of random noise. When random Poisson noise is added to the gene count martix, at increasing values of $\lambda$, a strategy based on optimization of modularity does return cell groups, whereas nSBM does not cluster cells when the noise exceeds a certain threshold (solid lines). The behaviour of the former approach follows the trend of the modularity (dashed lines), which is sustained at high levels independently from the level of noise.}\label{Figure_Random}
\end{figure}

\subsection*{Model hierarchy contains biological information}
The hierarchical model of cell groups implies that a relationship exists between groups. We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation. To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (\nameref{S3_Fig}A). After applying nSBM, we identify 27 groups at level 3 of the hierarchy (\nameref{S3_Fig}B), compared to the 24 using Leiden method at default resolution (\nameref{S4_Fig}). We found that the hierarchy proposed by our model is consistent with the developmental model (Fig. \ref{Figure2}). Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (\nameref{S4_Fig}); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.

In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
\end{figure}

\subsection*{Cell affinities can be used to evaluate cluster purity}
The computational framework underlying \emph{schist} calculates the model entropy, that is the amount of information required to describe a block configuration. Given that minimisation of such quantity can be used to perform model selection, it can be also used to evaluate the impact of modifying the assignment of a cell to a cluster. Once a model is minimised, \emph{schist} performs an exhaustive exploration of all model entropies resulting from moving all cells into all possible clusters. The differences in entropies could be interpreted as affinities of cells to given clusters. Such affinities are, in fact, probability values and could be used to evaluate the internal consistency of a given cell cluster.

To this end we calculate the entropy of the group-wise distribution of cell affinities, which is maximal when all cells have affinity equal to 1 for a given group. We tested this idea on four datasets recently published to benchmark single cell technologies in the Human Cell Atlas project \cite{mereu_2020}; in particular, we chose two technologies resulting in high quality data: Quartz-seq2 \cite{sasagawa_2018} and Chromium 10x v3 \cite{zheng_2017}, and two technologies resulting in more noisy data: MARS-seq \cite{jaitin_2014} and iCell8 \cite{goldstein_2017} (Fig. \ref{Figure3}).

Cluster consistency is not a measure of the data quality, in fact we identify low consistency groups in all datasets. High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper. Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells ( \nameref{S5_Fig}).

Similarly, we can use cell affinities to derive a stability parameter, a measure of the tendency for a cell to be stably associated to given clusters at all levels of the hierarchy. To this end, we first calculate the cell-wise entropy $H_{i,h}$ of cell affinity at each hierarchy level \emph{h}, then we define the stability as $S_i = 1 - \max(H_i)$. While we conceived this measure to identify and exclude cells with dubious assignment, we found that it may be more useful to assess the general data quality: the fraction of cells having $S>0.95$ was 0.783, 0.795, 0.831 and 0.855 for the iCELL8, MARS-seq, Chromium 10x and Quartz-seq2 technology respectively, in line with the evaluation on increasing performances of those platforms in \cite{mereu_2020}.

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
\caption[]{Analysis of cell cluster consistency. Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform. Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1. The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster. Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
\end{figure}

%\subsection*{Hierarchic model constrains differential gene tests}
%Current approaches to test for gene markers in scRNA-seq data rely on pairwise or one-vs-rest comparisons among cell groups. This strategy assumes that cell groups are independent which, in general, may not be true. A notable exception is the recently introduced treeclimbR \cite{huang_2020} which has been specifically developed to perform differential tests over a tree structure. We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.

\subsection*{Analysis of runtimes}
Minimisation of the nSBM is a process that may require a large amount of computational resources. The analysis of a relatively small scRNA-seq dataset, such as the ones in \cite{mereu_2020}, may require several minutes to be processed. This could be a serious limitation to the adoption of nSBM in the analysis of single cell data, especially because several parameters should be tested. To overcome this limitation, it was suggested to let a greedy merge-split MCMC algoritthm \cite{peixoto_2020} to explore the solutions and stop iterations when the difference in entropy is below a defined threshold. We tested this approach on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM) and compared to the default approach. Results are reported in Table \ref{Table1}. The merge-split algorithm greatly reduces the time needed to propose the final model. In addition, the partitions found are largely overlapping the ones found by the default approach.

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c  ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Minimize} & \textbf{Merge-split MCMC} & \textbf{Overlap} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{tian_2019} & 860 & 01:11 & 00:03 & 0.884 \\ 
 \hline
 Quartzseq \cite{mereu_2020} & 1266 & 00:31 & 00:02 & 0.726 \\ 
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 00:29 & 00:08 & 0.834 \\
 \hline
 Chromium 10X \cite{mereu_2020} & 1523 & 00:40 & 00:04 & 0.695 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 00:54 & 00:07 & 0.623 \\
 \hline
 Paul15 \cite{paul_2015} & 2730 & 02:37 & 00:10 & 0.575 \\ 
 \hline
 Planaria \cite{plass_2018} & 21612 & 13:12 & 03:29 & 0.589 \\
 \hline
% 4 & 545 & 18744 & 7560 \\
% \hline
% 5 & 88 & 788 & 6344 \\ [1ex] 
% \hline
\end{tabular}
\caption{Time required to minimise the nSBM using the default minimization method compared to the greedy merge-split MCMC. Times are expressed in mm:ss. Partition overlap measures concordance between the two models over the full hierarchy. Timing is the average after 3 initialisations.}
\label{Table1}
\end{table}

\section*{Conclusion}

Identification of cells sharing similar properties in single cell experiments is of paramount importance. A large number of approaches have been described, although the standardisation of analysis pipelines converged to methods that are based on modularity optimisation. We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques. The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution. In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble. Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity. Lastly, a statistical way to evaluate models is made available, allowing for reliable model selection. This last capability has the obvious advantage that the choice of parameters, hence the definition of cell clusters, could be conditioned to an evaluation metric which is robust and easy to understand (\emph{i.e.} the model entropy).

The major drawback of adopting this strategy is the substantial increase of runtimes. According to the developers of \emph{graph-tool}, runtimes are proportional to the number of edges in the neighbourhood graph and while it supports CPU-level parallelisation, a model minimisation is hundreds times slower than the extremely fast Leiden approach. Nevertheless, we show that a greedy merge-split MCMC algorithm can overcome this limitation, achieving performances that allow the usage of \emph{schist} on standard desktop hardware to analyse various single cell datasets.

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Degree distribution of \emph{k}NN graphs.} Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors. Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis). Both the parameters influence the sparseness of the graph.

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf nSBM performance after MCMC.}  Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth.

\paragraph*{S3 Fig.}
\label{S3_Fig}
{\bf Analysis of hematopoietic differentiation. }  (A) Low dimensional embedding of single cells colored by original cell type and pseudotime. (B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Non hierarchical clustering at different resolution.}  Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1. Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution \emph{r}=0.3 splits in groups 2 and 0 at resolution \emph{r}=0.2.

\paragraph*{S5 Fig.}
\label{S5_Fig}
{\bf Cluster consistency and cell type identification.} Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency. Dot size is proportional to the number of cells remaining after filtering.

\section*{Acknowledgments}
We would like to thank Tiago de Paula Peixoto (Central European University, ISI Foundation) and Giovanni Petri (ISI Foundation) for the discussions and the precious hints. We also would like to thank all people at COSR, in particular Giovanni Tonon and Paolo Provero.

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\begin{thebibliography}{10}

\bibitem{svensson_2018}
Svensson V, Vento-Tormo R, Teichmann SA.
\newblock Exponential scaling of single-cell {RNA}-seq in the past decade.
\newblock Nature Protocols. 2018;13(4):599--604.
\newblock doi:{10.1038/nprot.2017.149}.

\bibitem{guo_2018}
Guo J, Grow EJ, Mlcochova H, Maher GJ, Lindskog C, Nie X, et~al.
\newblock The adult human testis transcriptional cell atlas.
\newblock Cell Research. 2018;28(12):1141--1157.
\newblock doi:{10.1038/s41422-018-0099-2}.

\bibitem{ventotormo_2018}
Vento-Tormo R, Efremova M, Botting RA, Turco MY, Vento-Tormo M, Meyer KB,
  et~al.
\newblock Single-cell reconstruction of the early maternal-fetal interface in
  humans.
\newblock Nature. 2018;563(7731):347--353.
\newblock doi:{10.1038/s41586-018-0698-6}.

\bibitem{rozenblattrosen_2020}
Rozenblatt-Rosen O, Regev A, Oberdoerffer P, Nawy T, Hupalowska A, Rood JE,
  et~al.
\newblock The Human Tumor Atlas Network: Charting Tumor Transitions across
  Space and Time at Single-Cell Resolution.
\newblock Cell. 2020;181(2):236--249.
\newblock doi:{10.1016/j.cell.2020.03.053}.

\bibitem{tirosh_2016}
Tirosh I, Izar B, Prakadan SM, Wadsworth MH, Treacy D, Trombetta JJ, et~al.
\newblock Dissecting the multicellular ecosystem of metastatic melanoma by
  single-cell {RNA}-seq.
\newblock Science. 2016;352(6282):189--196.
\newblock doi:{10.1126/science.aad0501}.

\bibitem{patel_2014}
Patel AP, Tirosh I, Trombetta JJ, Shalek AK, Gillespie SM, Wakimoto H, et~al.
\newblock Single-cell {RNA}-seq highlights intratumoral heterogeneity in
  primary glioblastoma.
\newblock Science. 2014;344(6190):1396--1401.
\newblock doi:{10.1126/science.1254257}.

\bibitem{neftel_2019}
Neftel C, Laffy J, Filbin MG, Hara T, Shore ME, Rahme GJ, et~al.
\newblock An integrative model of cellular states, plasticity, and genetics for
  glioblastoma.
\newblock Cell. 2019;178(4):835--849.e21.
\newblock doi:{10.1016/j.cell.2019.06.024}.

\bibitem{rosenberg_2018}
Rosenberg AB, Roco CM, Muscat RA, Kuchina A, Sample P, Yao Z, et~al.
\newblock Single-cell profiling of the developing mouse brain and spinal cord
  with split-pool barcoding.
\newblock Science. 2018;360(6385):176--182.
\newblock doi:{10.1126/science.aam8999}.

\bibitem{wagner_2018}
Wagner DE, Weinreb C, Collins ZM, Briggs JA, Megason SG, Klein AM.
\newblock Single-cell mapping of gene expression landscapes and lineage in the
  zebrafish embryo.
\newblock Science. 2018;360(6392):981--987.
\newblock doi:{10.1126/science.aar4362}.

\bibitem{plass_2018}
Plass M, Solana J, Wolf FA, Ayoub S, Misios A, Glažar P, et~al.
\newblock Cell type atlas and lineage tree of a whole complex animal by
  single-cell transcriptomics.
\newblock Science. 2018;360(6391).
\newblock doi:{10.1126/science.aaq1723}.

\bibitem{regev_2017}
Regev A, Teichmann SA, Lander ES, Amit I, Benoist C, Birney E, et~al.
\newblock The human cell atlas.
\newblock eLife. 2017;6.
\newblock doi:{10.7554/{eLife}.27041}.

\bibitem{wang_2017}
Wang B, Zhu J, Pierson E, Ramazzotti D, Batzoglou S.
\newblock Visualization and analysis of single-cell {RNA}-seq data by
  kernel-based similarity learning.
\newblock Nature Methods. 2017;14(4):414--416.
\newblock doi:{10.1038/nmeth.4207}.

\bibitem{lin_2017}
Lin P, Troup M, Ho JWK.
\newblock {CIDR}: Ultrafast and accurate clustering through imputation for
  single-cell {RNA}-seq data.
\newblock Genome Biology. 2017;18(1):59.
\newblock doi:{10.1186/s13059-017-1188-0}.

\bibitem{huh_2020}
Huh R, Yang Y, Jiang Y, Shen Y, Li Y.
\newblock {SAME}-clustering: Single-cell Aggregated Clustering via Mixture
  Model Ensemble.
\newblock Nucleic Acids Research. 2020;48(1):86--95.
\newblock doi:{10.1093/nar/gkz959}.

\bibitem{kiselev_2017}
Kiselev VY, Kirschner K, Schaub MT, Andrews T, Yiu A, Chandra T, et~al.
\newblock {SC3}: consensus clustering of single-cell {RNA}-seq data.
\newblock Nature Methods. 2017;14(5):483--486.
\newblock doi:{10.1038/nmeth.4236}.

\bibitem{li_2020}
Li X, Wang K, Lyu Y, Pan H, Zhang J, Stambolian D, et~al.
\newblock Deep learning enables accurate clustering with batch effect removal
  in single-cell {RNA}-seq analysis.
\newblock Nature Communications. 2020;11(1):2338.
\newblock doi:{10.1038/s41467-020-15851-3}.

\bibitem{krzak_2019}
Krzak M, Raykov Y, Boukouvalas A, Cutillo L, Angelini C.
\newblock Benchmark and Parameter Sensitivity Analysis of Single-Cell {RNA}
  Sequencing Clustering Methods.
\newblock Frontiers in genetics. 2019;10:1253.
\newblock doi:{10.3389/fgene.2019.01253}.

\bibitem{kiselev_2019}
Kiselev VY, Andrews TS, Hemberg M.
\newblock Challenges in unsupervised clustering of single-cell {RNA}-seq data.
\newblock Nature Reviews Genetics. 2019;20(5):273--282.
\newblock doi:{10.1038/s41576-018-0088-9}.

\bibitem{du_2018}
{Duò} A, Robinson MD, Soneson C.
\newblock A systematic performance evaluation of clustering methods for
  single-cell {RNA}-seq data.
\newblock F1000Research. 2018;7:1141.
\newblock doi:{10.12688/f1000research.15666.2}.

\bibitem{butler_2018}
Butler A, Hoffman P, Smibert P, Papalexi E, Satija R.
\newblock Integrating single-cell transcriptomic data across different
  conditions, technologies, and species.
\newblock Nature Biotechnology. 2018;36(5):411--420.
\newblock doi:{10.1038/nbt.4096}.

\bibitem{wolf_2018}
Wolf FA, Angerer P, Theis FJ.
\newblock {SCANPY}: large-scale single-cell gene expression data analysis.
\newblock Genome Biology. 2018;19(1):15.
\newblock doi:{10.1186/s13059-017-1382-0}.

\bibitem{blondel_2008}
Blondel VD, Guillaume JL, Lambiotte R, Lefebvre E.
\newblock Fast unfolding of communities in large networks.
\newblock Journal of Statistical Mechanics: Theory and Experiment.
  2008;2008(10):P10008.
\newblock doi:{10.1088/1742-5468/2008/10/P10008}.

\bibitem{traag_2019}
Traag VA, Waltman L, van Eck NJ.
\newblock From Louvain to Leiden: guaranteeing well-connected communities.
\newblock Scientific Reports. 2019;9(1):5233.
\newblock doi:{10.1038/s41598-019-41695-z}.

\bibitem{levine_2015}
Levine JH, Simonds EF, Bendall SC, Davis KL, Amir EaD, Tadmor MD, et~al.
\newblock Data-Driven Phenotypic Dissection of {AML} Reveals Progenitor-like
  Cells that Correlate with Prognosis.
\newblock Cell. 2015;162(1):184--197.
\newblock doi:{10.1016/j.cell.2015.05.047}.

\bibitem{newman_2004}
Newman MEJ, Girvan M.
\newblock Finding and evaluating community structure in networks.
\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
  2004;69(2 Pt 2):026113.
\newblock doi:{10.1103/{PhysRevE}.69.026113}.

\bibitem{traag_2011}
Traag VA, Van~Dooren P, Nesterov Y.
\newblock Narrow scope for resolution-limit-free community detection.
\newblock Physical Review E. 2011;84(1).
\newblock doi:{10.1103/{PhysRevE}.84.016114}.

\bibitem{reichardt_2006}
Reichardt J, Bornholdt S.
\newblock Statistical mechanics of community detection.
\newblock Physical Review E. 2006;74(1).
\newblock doi:{10.1103/{PhysRevE}.74.016110}.

\bibitem{lhnemann_2020}
{Lähnemann} D, {Köster} J, Szczurek E, {McCarthy} DJ, Hicks SC, Robinson MD,
  et~al.
\newblock Eleven grand challenges in single-cell data science.
\newblock Genome Biology. 2020;21(1):31.
\newblock doi:{10.1186/s13059-020-1926-6}.

\bibitem{fortunato_2007}
Fortunato S, {Barthélemy} M.
\newblock Resolution limit in community detection.
\newblock Proceedings of the National Academy of Sciences of the United States
  of America. 2007;104(1):36--41.
\newblock doi:{10.1073/pnas.0605965104}.

\bibitem{guimer_2004}
{Guimerà} R, Sales-Pardo M, Amaral LAN.
\newblock Modularity from fluctuations in random graphs and complex networks.
\newblock Physical Review E. 2004;70(2).
\newblock doi:{10.1103/{PhysRevE}.70.025101}.

\bibitem{baran_2019}
Baran Y, Bercovich A, Sebe-Pedros A, Lubling Y, Giladi A, Chomsky E, et~al.
\newblock {MetaCell}: analysis of single-cell {RNA}-seq data using K-nn graph
  partitions.
\newblock Genome Biology. 2019;20(1):206.
\newblock doi:{10.1186/s13059-019-1812-2}.

\bibitem{xu_2015}
Xu C, Su Z.
\newblock Identification of cell types from single-cell transcriptomes using a
  novel clustering method.
\newblock Bioinformatics. 2015;31(12):1974--1980.
\newblock doi:{10.1093/bioinformatics/btv088}.

\bibitem{miao_2020}
Miao Z, Moreno P, Huang N, Papatheodorou I, Brazma A, Teichmann SA.
\newblock Putative cell type discovery from single-cell gene expression data.
\newblock Nature Methods. 2020;17(6):621--628.
\newblock doi:{10.1038/s41592-020-0825-9}.

\bibitem{holland_1983}
Holland PW, Laskey KB, Leinhardt S.
\newblock Stochastic blockmodels: First steps.
\newblock Social networks. 1983;5(2):109--137.
\newblock doi:{10.1016/0378-8733(83)90021-7}.

\bibitem{peixoto_2017}
Peixoto TP.
\newblock Nonparametric Bayesian inference of the microcanonical stochastic
  block model.
\newblock Physical review E. 2017;95(1-1):012317.
\newblock doi:{10.1103/{PhysRevE}.95.012317}.

\bibitem{karrer_2011}
Karrer B, Newman MEJ.
\newblock Stochastic blockmodels and community structure in networks.
\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
  2011;83(1 Pt 2):016107.
\newblock doi:{10.1103/{PhysRevE}.83.016107}.

\bibitem{peixoto_2013}
Peixoto TP.
\newblock Parsimonious module inference in large networks.
\newblock Physical Review Letters. 2013;110(14):148701.
\newblock doi:{10.1103/{PhysRevLett}.110.148701}.

\bibitem{peixoto_2014}
Peixoto TP.
\newblock Efficient Monte Carlo and greedy heuristic for the inference of
  stochastic block models.
\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
  2014;89(1):012804.
\newblock doi:{10.1103/{PhysRevE}.89.012804}.

\bibitem{peixoto_2014_h}
Peixoto TP.
\newblock Hierarchical Block Structures and High-Resolution Model Selection in
  Large Networks.
\newblock Physical Review X. 2014;4(1):011047.
\newblock doi:{10.1103/{PhysRevX}.4.011047}.

\bibitem{tian_2019}
Tian L, Dong X, Freytag S, Lê~Cao KA, Su S, {JalalAbadi} A, et~al.
\newblock Benchmarking single cell {RNA}-sequencing analysis pipelines using
  mixture control experiments.
\newblock Nature Methods. 2019;16(6):479--487.
\newblock doi:{10.1038/s41592-019-0425-8}.

\bibitem{kang_2018}
Kang HM, Subramaniam M, Targ S, Nguyen M, Maliskova L, {McCarthy} E, et~al.
\newblock Multiplexed droplet single-cell {RNA}-sequencing using natural
  genetic variation.
\newblock Nature Biotechnology. 2018;36(1):89--94.
\newblock doi:{10.1038/nbt.4042}.

\bibitem{mereu_2020}
Mereu E, Lafzi A, Moutinho C, Ziegenhain C, {McCarthy} DJ, {Álvarez-Varela} A,
  et~al.
\newblock Benchmarking single-cell {RNA}-sequencing protocols for cell atlas
  projects.
\newblock Nature Biotechnology. 2020;38(6):747--755.
\newblock doi:{10.1038/s41587-020-0469-4}.

\bibitem{sasagawa_2018}
Sasagawa Y, Danno H, Takada H, Ebisawa M, Tanaka K, Hayashi T, et~al.
\newblock Quartz-Seq2: a high-throughput single-cell {RNA}-sequencing method
  that effectively uses limited sequence reads.
\newblock Genome Biology. 2018;19(1):29.
\newblock doi:{10.1186/s13059-018-1407-3}.

\bibitem{zheng_2017}
Zheng GXY, Terry JM, Belgrader P, Ryvkin P, Bent ZW, Wilson R, et~al.
\newblock Massively parallel digital transcriptional profiling of single cells.
\newblock Nature Communications. 2017;8:14049.
\newblock doi:{10.1038/ncomms14049}.

\bibitem{jaitin_2014}
Jaitin DA, Kenigsberg E, Keren-Shaul H, Elefant N, Paul F, Zaretsky I, et~al.
\newblock Massively parallel single-cell {RNA}-seq for marker-free
  decomposition of tissues into cell types.
\newblock Science. 2014;343(6172):776--779.
\newblock doi:{10.1126/science.1247651}.

\bibitem{goldstein_2017}
Goldstein LD, Chen YJJ, Dunne J, Mir A, Hubschle H, Guillory J, et~al.
\newblock Massively parallel nanowell-based single-cell gene expression
  profiling.
\newblock {BMC} Genomics. 2017;18(1):519.
\newblock doi:{10.1186/s12864-017-3893-1}.

\bibitem{paul_2015}
Paul F, Arkin Y, Giladi A, Jaitin DA, Kenigsberg E, Keren-Shaul H, et~al.
\newblock Transcriptional heterogeneity and lineage commitment in myeloid
  progenitors.
\newblock Cell. 2015;163(7):1663--1677.
\newblock doi:{10.1016/j.cell.2015.11.013}.

\bibitem{wolf_2019}
Wolf FA, Hamey FK, Plass M, Solana J, Dahlin JS, {Göttgens} B, et~al.
\newblock {PAGA}: graph abstraction reconciles clustering with trajectory
  inference through a topology preserving map of single cells.
\newblock Genome Biology. 2019;20(1):59.
\newblock doi:{10.1186/s13059-019-1663-x}.

\bibitem{peixoto_2020}
Peixoto TP.
\newblock Merge-split Markov chain Monte Carlo for community detection.
\newblock arXiv. 2020;doi:{arXiv:2003.07070}.

\end{thebibliography}

\end{document}

### 2020-11-16 manuscript.tex
initial import of old files

--- 2020-11-16 PlosCompBio/plos_latex_template.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
+\documentclass[10pt]{article}
+\usepackage[top=20mm, bottom=20mm, left=20mm, right=20mm]{geometry}
+\geometry{a4paper}
+\geometry{portrait}
+\usepackage[utf8]{inputenc}	% Comment out for xelatex; must use with pdflatex
+\usepackage[T1]{fontenc}
+\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+%\usepackage{fontspec}
+\usepackage{epstopdf}
+\usepackage{float}
+\usepackage[hang]{footmisc}
+\usepackage{endnotes}
+\usepackage{listingsutf8}
+%\restylefloat{table}
+\usepackage{xcolor}
+\usepackage{colortbl}
+%\usepackage{booktabs}
+\usepackage{tabu}
+\usepackage{caption}
+\usepackage{textcomp}
+\usepackage{textgreek}
+\usepackage{subcaption}
+\usepackage{hyperref}
+\usepackage[normalem]{ulem}
+\usepackage{authblk}
+\usepackage[numbers]{natbib}
+\bibliographystyle{plos2015}
 
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\makeatletter
+\def\blx@maxline{77}
+\makeatother
 
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
+\captionsetup[figure]{labelfont=sc}
+\captionsetup[table]{labelfont=sc}
+\captionsetup[tabu]{labelfont=sc}
 
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
+\DeclareCaptionType{MPEquation}[][List of equations]
+\captionsetup[MPEquation]{labelformat=empty}
 
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
+\DeclareCaptionType{MPListing}[][List of listings]
+\captionsetup[MPListing]{labelformat=empty}
 
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
+\usepackage{fancyhdr}
+\setlength{\headheight}{16pt}
+\pagestyle{fancyplain}
+\fancyhf{}
+%\fancypagestyle{plain}{}
+\cfoot[\thepage]{\thepage}
 
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
+\usepackage{stringenc}
+\usepackage{pdfescape}
 
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
+% Helper for debugging pdflatex errors induced by invalid UTF8 characters:
+% output the actual character code causing the issue.
 
-% line numbers
-\usepackage[right]{lineno}
+\makeatletter
+\renewcommand*{\UTFviii@defined}[1]{%
+\ifx#1\relax
+\begingroup
+% Remove prefix "\u8:"
+\def\x##1:{}%
+% Extract Unicode char from command name
+% (utf8.def does not support surrogates)
+\edef\x{\expandafter\x\string#1}%
+\StringEncodingConvert\x\x{utf8}{utf16be}% convert to UTF-16BE
+% Hexadecimal representation
+\EdefEscapeHex\x\x
+% Enhanced error message
+\PackageError{inputenc}{Unicode\space char\space \string#1\space
+(U+\x)\MessageBreak
+not\space set\space up\space
+for\space use\space with\space LaTeX}\@eha
+\endgroup
+\else\expandafter
+#1%
+\fi
+}
+\makeatother
 
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
+% Make multi-paragraph footnotes align nicely on the left
+\setlength\footnotemargin{10pt}
 
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
+% Do not display an automatic Notes heading for endnotes, as they are contained
+% by a Manuscripts section with a heading that has possibly been edited by the user
+\def\enoteheading{}
 
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
+% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
+\renewcommand{\thefootnote}{\arabic{footnote}}
+\renewcommand{\theendnote}{\arabic{endnote}}
+% Adjust endnotes to use normal-sized numbering and text
+\renewcommand{\enotesize}{\normalsize}
+\renewcommand\enoteformat{%
+  \raggedright
+  \leftskip=1.8em
+  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
 }
 
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
+\begin{document}
+\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
+\author[1,2]{Leonardo Morelli}
+\author[1,3]{Valentina Giansanti}
+\author[1]{Davide Cittaro}
+\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
+\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
+\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
+\maketitle
 
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
+\section*{Abstract}
 
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Nested stochastic block models applied to the analysis of single cell data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Leonardo Morelli\textsuperscript{1,2},
-Valentina Giansanti\textsuperscript{1, 3},
-Davide Cittaro\textsuperscript{1*},
-\\
-\bigskip
-\textbf{1} Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy
-\\
-\textbf{2} Universit\`{a} Vita-Salute San Raffaele, Milan, Italy
-\\
-\textbf{3} Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy
-\\
-\bigskip
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
-
-% Current address notes
-%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-%\dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Membership list can be found in the Acknowledgments section.
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* cittaro.davide@hsr.it
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
 Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system.
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-Identification of cell types is a key step of many single cell experiments.
-Many of the approaches to achieve this goal converge on the analysis of the graph representing cell-wise similarity by optimisation of modularity.
-While fast, this method does not provide a robust estimation of the number of groups, and it leaves large room of arbitrariness in the choice of appropriate resolution.
-To overcome these limitations, we propose a strategy based on nested stochastic block models.
-We show our method is accurate and provides a rich description of cell groups and their relationships.
+\section*{Background}
 
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
 Transcriptome analysis at single cell level by RNA sequencing (scRNA-seq) is a technology growing in popularity and applications \cite{svensson_2018}.
@@
 
-%\begin{MPEquation}[!ht]
+\begin{MPEquation}[!ht]
 \begin{equation}
@@
 \label{MPEquationElement:820CB015-E51E-4C5A-BD9C-70E667B8F73B}
-%\end{MPEquation}
+\end{MPEquation}
 and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length
 
-%\begin{MPEquation}[!ht]
+\begin{MPEquation}[!ht]
 \begin{equation}
@@
 \label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
-%\end{MPEquation}
+\end{MPEquation}
 so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
@@
 
-\section*{Materials and methods}
-%\subsection*{Analysis of Random data}
-%Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function.
-The random \emph{k}NN graph was obtained shuffling the node labels of each edge.
-nSBM was performed with 3 initialisations using merge-split MCMC approach.
-UMAP embedding was recomputed after randomisation using the shuffled graph.
+\section*{Results}
 
-\subsection*{Analysis of cell mixtures}
-
-Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}).
-Data were analysed using scanpy v1.4.6 \cite{wolf_2018}.
-Cells with less than 200 genes were excluded, as genes detected in less than 3 cells.
-Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis.
-Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out.
-nSBM was initialised three times.
-Analysis was performed at level 2 of the nSBM hierarchy.
-Random Poisson noise was generated at given $\lambda$ (range: 0 - 1000) and added to the initial count matrix. Analysis was performed at level 2 of the nSBM hierarchy.
-
-\subsection*{Analysis of hematopoietic differentiation}
-
-Data were retrieved using scanpy's built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbours and diffmap as embedding.
-Gene signatures were calculated using the following gene lists
-\begin{itemize}
-\item Erythroids: Gata1, Klf1, Epor, Gypa, Hba-a2, Hba-a1, Spi1
-\item Neutrophils, Elane, Cebpe, Ctsg, Mpo, Gfi1
-\item Monocytes, Irf8, Csf1r, Ctsg, Mpo
-\end{itemize}
-
-nSBM was completed with 3 initialisations
-
-\subsection*{Analysis of cluster consistency}
-
-Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8).
-Data were processed according to the methods in the original paper \cite{mereu_2020}.
-Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded.
-Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses.
-Genes that were expressed in less than five cells were removed.
-Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5.
-Neighbourhood graph was built using 30 principal components and 20 neighbours.
-nSBM was completed with 3 initialisations.
-
-% Results and Discussion can be combined.
 \subsection*{Overview of \emph{schist}}
@@
 
-%\subsection*{nSBM does not return artificial cell groupings}
-%One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling.
+\subsection*{nSBM does not return artificial cell groupings}
+One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling.
 In this sense, the number of groups identified strictly mirrors the amount of information contained in the data.
@@
 
-%This experiment is a deliberate extreme case.
-The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information.
-Nevertheless, real-world data may include an unknown amount of random noise.
-Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.
-
-%\begin{figure}[!h]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
-%\caption[]{nSBM results for randomised data.
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
+\caption[]{nSBM results for randomised data.
 (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right).
@@
 In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
-%\end{figure}
+\end{figure}
+
+This experiment is a deliberate extreme case.
+The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information.
+Nevertheless, real-world data may include an unknown amount of random noise.
+Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.
 
@@
 Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours.
-If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (\nameref{S1_Fig}).
-Running MCMC algorithm recovers the performances of the majority of the configurations (\nameref{S2_Fig}).
+If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (Fig.
+S1).
+Running MCMC algorithm recovers the performances of the majority of the configurations (Fig.
+S2).
 
-Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig.
-\ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$).
-In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3.
-Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them.
-It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
-As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig.
-\ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.
-
-It is worth noting that since nSBM reflects the amount of information contained in the data, it does not return clusters when data are noisy.
-To demonstrate this property, we added increasing levels of random Poisson noise to the count matrix (Fig.\ref{Figure_Random}).
-The number of groups reported by nSBM decreases with increasing amount of noise and it is 1 (\emph{i.e.} no clusters) at $\lambda=200$, whereas a standard approach overfits data.
-
-These observations show that nSBM is suitable for accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter and with the possibility to identify rare cell types in larger populations.
-Moreover, nSBM avoids excessive clustering of data in presence of noisy measurements.
-
-\begin{figure}[!h]
+\begin{figure}[H]
 \centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
 \caption[]{\emph{schist} applied to scRNA-seq mixology data.
@@
 
-\begin{figure}[!h]
-\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Tian_Random.png}
-\caption[]{Identification of cell groups at different level of random noise.
-When random Poisson noise is added to the gene count martix, at increasing values of $\lambda$, a strategy based on optimization of modularity does return cell groups, whereas nSBM does not cluster cells when the noise exceeds a certain threshold (solid lines).
-The behaviour of the former approach follows the trend of the modularity (dashed lines), which is sustained at high levels independently from the level of noise.}\label{Figure_Random}
-\end{figure}
+Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig.
+\ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$).
+In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3.
+Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them.
+It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
+As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig.
+\ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.
 
-\subsection*{Model hierarchy contains biological information}
-The hierarchical model of cell groups implies that a relationship exists between groups.
-We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation.
-To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}.
-Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (\nameref{S3_Fig}A).
-After applying nSBM, we identify 27 groups at level 3 of the hierarchy (\nameref{S3_Fig}B), compared to the 24 using Leiden method at default resolution (\nameref{S4_Fig}).
-We found that the hierarchy proposed by our model is consistent with the developmental model (Fig.
-\ref{Figure2}).
-Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (\nameref{S4_Fig}); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.
-
-In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.
-
-\begin{figure}[!h]
-\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
-\caption[]{Analysis of hematopoietic differentiation.
-Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy.
-Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A).
-In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
-\end{figure}
+These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter.
+These data also hint at the possibility to identify rare cell types in larger populations.
 
@@
 
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
+\caption[]{Analysis of cell cluster consistency.
+Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform.
+Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1.
+The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster.
+Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
+\end{figure}
+
 To this end we calculate the entropy of the group-wise distribution of cell affinities, which is maximal when all cells have affinity equal to 1 for a given group.
@@
 High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper.
-Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells ( \nameref{S5_Fig}).
+Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells (Fig.
+S5).
 
@@
 
-\begin{figure}[!h]
+\subsection*{Model hierarchy contains biological information}
+The hierarchical model of cell groups implies that a relationship exists between groups.
+We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation.
+To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}.
+Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig.
+S3A).
+After applying nSBM, we identify 27 groups at level 3 of the hierarchy (Fig.
+S3B), compared to the 24 using Leiden method at default resolution (Fig.
+S4).
+We found that the hierarchy proposed by our model is consistent with the developmental model (Fig.
+\ref{Figure2}).
+Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (Fig.
+S4); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.
+
+\begin{figure}[H]
 \centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
-\caption[]{Analysis of cell cluster consistency.
-Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform.
-Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1.
-The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster.
-Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
+\caption[]{Analysis of hematopoietic differentiation.
+Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy.
+Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A).
+In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
 \end{figure}
+
+In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.
 
@@
 
-\section*{Conclusion}
+\section*{Conclusions}
 
@@
 
-\section*{Supporting information}
+\section*{Materials and Methods}
 
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Degree distribution of \emph{k}NN graphs.} Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors.
-Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis).
-Both the parameters influence the sparseness of the graph.
+\subsection*{Analysis of Random data}
+Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function.
+The random \emph{k}NN graph was obtained shuffling the node labels of each edge.
+nSBM was performed with 3 initialisations using merge-split MCMC approach.
+UMAP embedding was recomputed after randomisation using the shuffled graph.
 
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf nSBM performance after MCMC.}  Adjusted Rand Index for different \emph{k}NN graphs after MCMC run.
-Maximal ARI over all hierarchy level is shown.
-Darker color indicates higher concordance with the ground truth.
+\subsection*{Analysis of cell mixtures}
 
-\paragraph*{S3 Fig.}
-\label{S3_Fig}
-{\bf Analysis of hematopoietic differentiation.
-}  (A) Low dimensional embedding of single cells colored by original cell type and pseudotime.
-(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.
+Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}).
+Data were analysed using scanpy v1.4.6 \cite{wolf_2018}.
+Cells with less than 200 genes were excluded, as genes detected in less than 3 cells.
+Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis.
+Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out.
+nSBM was initialised three times
 
-\paragraph*{S4 Fig.}
-\label{S4_Fig}
-{\bf Non hierarchical clustering at different resolution.}  Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution \emph{r}=0.3 splits in groups 2 and 0 at resolution \emph{r}=0.2.
+\subsection*{Analysis of hematopoietic differentiation}
 
-\paragraph*{S5 Fig.}
-\label{S5_Fig}
-{\bf Cluster consistency and cell type identification.} Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency.
-Dot size is proportional to the number of cells remaining after filtering.
+Data were retrieved using scanpy's built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbours and diffmap as embedding.
+Gene signatures were calculated using the following gene lists
+\begin{itemize}
+\item Erythroids: Gata1, Klf1, Epor, Gypa, Hba-a2, Hba-a1, Spi1
+\item Neutrophils, Elane, Cebpe, Ctsg, Mpo, Gfi1
+\item Monocytes, Irf8, Csf1r, Ctsg, Mpo
+\end{itemize}
 
-\section*{Acknowledgments}
+nSBM was completed with 3 initialisations
+
+\subsection*{Analysis of cluster consistency}
+
+Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8).
+Data were processed according to the methods in the original paper \cite{mereu_2020}.
+Briefly, cells with less than 10,000 total number of reads as well as the cells having less than 65\% of the reads mapped to their reference genome were discarded.
+Cells in the 95th percentile of the number of genes/cell and those having less than 25\% mitochondrial gene content were included in the downstream analyses.
+Genes that were expressed in less than five cells were removed.
+Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5.
+Neighbourhood graph was built using 30 principal components and 20 neighbours.
+nSBM was completed with 3 initialisations.
+
+\section*{Acknowledgements}
+
 We would like to thank Tiago de Paula Peixoto (Central European University, ISI Foundation) and Giovanni Petri (ISI Foundation) for the discussions and the precious hints.
 We also would like to thank all people at COSR, in particular Giovanni Tonon and Paolo Provero.
+This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC
 
-\nolinenumbers
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-% 
-\begin{thebibliography}{10}
-
-\bibitem{svensson_2018}
-Svensson V, Vento-Tormo R, Teichmann SA.
-\newblock Exponential scaling of single-cell {RNA}-seq in the past decade.
-\newblock Nature Protocols. 2018;13(4):599--604.
-\newblock doi:{10.1038/nprot.2017.149}.
-
-\bibitem{guo_2018}
-Guo J, Grow EJ, Mlcochova H, Maher GJ, Lindskog C, Nie X, et~al.
-\newblock The adult human testis transcriptional cell atlas.
-\newblock Cell Research. 2018;28(12):1141--1157.
-\newblock doi:{10.1038/s41422-018-0099-2}.
-
-\bibitem{ventotormo_2018}
-Vento-Tormo R, Efremova M, Botting RA, Turco MY, Vento-Tormo M, Meyer KB,
-  et~al.
-\newblock Single-cell reconstruction of the early maternal-fetal interface in
-  humans.
-\newblock Nature. 2018;563(7731):347--353.
-\newblock doi:{10.1038/s41586-018-0698-6}.
-
-\bibitem{rozenblattrosen_2020}
-Rozenblatt-Rosen O, Regev A, Oberdoerffer P, Nawy T, Hupalowska A, Rood JE,
-  et~al.
-\newblock The Human Tumor Atlas Network: Charting Tumor Transitions across
-  Space and Time at Single-Cell Resolution.
-\newblock Cell. 2020;181(2):236--249.
-\newblock doi:{10.1016/j.cell.2020.03.053}.
-
-\bibitem{tirosh_2016}
-Tirosh I, Izar B, Prakadan SM, Wadsworth MH, Treacy D, Trombetta JJ, et~al.
-\newblock Dissecting the multicellular ecosystem of metastatic melanoma by
-  single-cell {RNA}-seq.
-\newblock Science. 2016;352(6282):189--196.
-\newblock doi:{10.1126/science.aad0501}.
-
-\bibitem{patel_2014}
-Patel AP, Tirosh I, Trombetta JJ, Shalek AK, Gillespie SM, Wakimoto H, et~al.
-\newblock Single-cell {RNA}-seq highlights intratumoral heterogeneity in
-  primary glioblastoma.
-\newblock Science. 2014;344(6190):1396--1401.
-\newblock doi:{10.1126/science.1254257}.
-
-\bibitem{neftel_2019}
-Neftel C, Laffy J, Filbin MG, Hara T, Shore ME, Rahme GJ, et~al.
-\newblock An integrative model of cellular states, plasticity, and genetics for
-  glioblastoma.
-\newblock Cell. 2019;178(4):835--849.e21.
-\newblock doi:{10.1016/j.cell.2019.06.024}.
-
-\bibitem{rosenberg_2018}
-Rosenberg AB, Roco CM, Muscat RA, Kuchina A, Sample P, Yao Z, et~al.
-\newblock Single-cell profiling of the developing mouse brain and spinal cord
-  with split-pool barcoding.
-\newblock Science. 2018;360(6385):176--182.
-\newblock doi:{10.1126/science.aam8999}.
-
-\bibitem{wagner_2018}
-Wagner DE, Weinreb C, Collins ZM, Briggs JA, Megason SG, Klein AM.
-\newblock Single-cell mapping of gene expression landscapes and lineage in the
-  zebrafish embryo.
-\newblock Science. 2018;360(6392):981--987.
-\newblock doi:{10.1126/science.aar4362}.
-
-\bibitem{plass_2018}
-Plass M, Solana J, Wolf FA, Ayoub S, Misios A, Glažar P, et~al.
-\newblock Cell type atlas and lineage tree of a whole complex animal by
-  single-cell transcriptomics.
-\newblock Science. 2018;360(6391).
-\newblock doi:{10.1126/science.aaq1723}.
-
-\bibitem{regev_2017}
-Regev A, Teichmann SA, Lander ES, Amit I, Benoist C, Birney E, et~al.
-\newblock The human cell atlas.
-\newblock eLife. 2017;6.
-\newblock doi:{10.7554/{eLife}.27041}.
-
-\bibitem{wang_2017}
-Wang B, Zhu J, Pierson E, Ramazzotti D, Batzoglou S.
-\newblock Visualization and analysis of single-cell {RNA}-seq data by
-  kernel-based similarity learning.
-\newblock Nature Methods. 2017;14(4):414--416.
-\newblock doi:{10.1038/nmeth.4207}.
-
-\bibitem{lin_2017}
-Lin P, Troup M, Ho JWK.
-\newblock {CIDR}: Ultrafast and accurate clustering through imputation for
-  single-cell {RNA}-seq data.
-\newblock Genome Biology. 2017;18(1):59.
-\newblock doi:{10.1186/s13059-017-1188-0}.
-
-\bibitem{huh_2020}
-Huh R, Yang Y, Jiang Y, Shen Y, Li Y.
-\newblock {SAME}-clustering: Single-cell Aggregated Clustering via Mixture
-  Model Ensemble.
-\newblock Nucleic Acids Research. 2020;48(1):86--95.
-\newblock doi:{10.1093/nar/gkz959}.
-
-\bibitem{kiselev_2017}
-Kiselev VY, Kirschner K, Schaub MT, Andrews T, Yiu A, Chandra T, et~al.
-\newblock {SC3}: consensus clustering of single-cell {RNA}-seq data.
-\newblock Nature Methods. 2017;14(5):483--486.
-\newblock doi:{10.1038/nmeth.4236}.
-
-\bibitem{li_2020}
-Li X, Wang K, Lyu Y, Pan H, Zhang J, Stambolian D, et~al.
-\newblock Deep learning enables accurate clustering with batch effect removal
-  in single-cell {RNA}-seq analysis.
-\newblock Nature Communications. 2020;11(1):2338.
-\newblock doi:{10.1038/s41467-020-15851-3}.
-
-\bibitem{krzak_2019}
-Krzak M, Raykov Y, Boukouvalas A, Cutillo L, Angelini C.
-\newblock Benchmark and Parameter Sensitivity Analysis of Single-Cell {RNA}
-  Sequencing Clustering Methods.
-\newblock Frontiers in genetics. 2019;10:1253.
-\newblock doi:{10.3389/fgene.2019.01253}.
-
-\bibitem{kiselev_2019}
-Kiselev VY, Andrews TS, Hemberg M.
-\newblock Challenges in unsupervised clustering of single-cell {RNA}-seq data.
-\newblock Nature Reviews Genetics. 2019;20(5):273--282.
-\newblock doi:{10.1038/s41576-018-0088-9}.
-
-\bibitem{du_2018}
-{Duò} A, Robinson MD, Soneson C.
-\newblock A systematic performance evaluation of clustering methods for
-  single-cell {RNA}-seq data.
-\newblock F1000Research. 2018;7:1141.
-\newblock doi:{10.12688/f1000research.15666.2}.
-
-\bibitem{butler_2018}
-Butler A, Hoffman P, Smibert P, Papalexi E, Satija R.
-\newblock Integrating single-cell transcriptomic data across different
-  conditions, technologies, and species.
-\newblock Nature Biotechnology. 2018;36(5):411--420.
-\newblock doi:{10.1038/nbt.4096}.
-
-\bibitem{wolf_2018}
-Wolf FA, Angerer P, Theis FJ.
-\newblock {SCANPY}: large-scale single-cell gene expression data analysis.
-\newblock Genome Biology. 2018;19(1):15.
-\newblock doi:{10.1186/s13059-017-1382-0}.
-
-\bibitem{blondel_2008}
-Blondel VD, Guillaume JL, Lambiotte R, Lefebvre E.
-\newblock Fast unfolding of communities in large networks.
-\newblock Journal of Statistical Mechanics: Theory and Experiment.
-  2008;2008(10):P10008.
-\newblock doi:{10.1088/1742-5468/2008/10/P10008}.
-
-\bibitem{traag_2019}
-Traag VA, Waltman L, van Eck NJ.
-\newblock From Louvain to Leiden: guaranteeing well-connected communities.
-\newblock Scientific Reports. 2019;9(1):5233.
-\newblock doi:{10.1038/s41598-019-41695-z}.
-
-\bibitem{levine_2015}
-Levine JH, Simonds EF, Bendall SC, Davis KL, Amir EaD, Tadmor MD, et~al.
-\newblock Data-Driven Phenotypic Dissection of {AML} Reveals Progenitor-like
-  Cells that Correlate with Prognosis.
-\newblock Cell. 2015;162(1):184--197.
-\newblock doi:{10.1016/j.cell.2015.05.047}.
-
-\bibitem{newman_2004}
-Newman MEJ, Girvan M.
-\newblock Finding and evaluating community structure in networks.
-\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
-  2004;69(2 Pt 2):026113.
-\newblock doi:{10.1103/{PhysRevE}.69.026113}.
-
-\bibitem{traag_2011}
-Traag VA, Van~Dooren P, Nesterov Y.
-\newblock Narrow scope for resolution-limit-free community detection.
-\newblock Physical Review E. 2011;84(1).
-\newblock doi:{10.1103/{PhysRevE}.84.016114}.
-
-\bibitem{reichardt_2006}
-Reichardt J, Bornholdt S.
-\newblock Statistical mechanics of community detection.
-\newblock Physical Review E. 2006;74(1).
-\newblock doi:{10.1103/{PhysRevE}.74.016110}.
-
-\bibitem{lhnemann_2020}
-{Lähnemann} D, {Köster} J, Szczurek E, {McCarthy} DJ, Hicks SC, Robinson MD,
-  et~al.
-\newblock Eleven grand challenges in single-cell data science.
-\newblock Genome Biology. 2020;21(1):31.
-\newblock doi:{10.1186/s13059-020-1926-6}.
-
-\bibitem{fortunato_2007}
-Fortunato S, {Barthélemy} M.
-\newblock Resolution limit in community detection.
-\newblock Proceedings of the National Academy of Sciences of the United States
-  of America. 2007;104(1):36--41.
-\newblock doi:{10.1073/pnas.0605965104}.
-
-\bibitem{guimer_2004}
-{Guimerà} R, Sales-Pardo M, Amaral LAN.
-\newblock Modularity from fluctuations in random graphs and complex networks.
-\newblock Physical Review E. 2004;70(2).
-\newblock doi:{10.1103/{PhysRevE}.70.025101}.
-
-\bibitem{baran_2019}
-Baran Y, Bercovich A, Sebe-Pedros A, Lubling Y, Giladi A, Chomsky E, et~al.
-\newblock {MetaCell}: analysis of single-cell {RNA}-seq data using K-nn graph
-  partitions.
-\newblock Genome Biology. 2019;20(1):206.
-\newblock doi:{10.1186/s13059-019-1812-2}.
-
-\bibitem{xu_2015}
-Xu C, Su Z.
-\newblock Identification of cell types from single-cell transcriptomes using a
-  novel clustering method.
-\newblock Bioinformatics. 2015;31(12):1974--1980.
-\newblock doi:{10.1093/bioinformatics/btv088}.
-
-\bibitem{miao_2020}
-Miao Z, Moreno P, Huang N, Papatheodorou I, Brazma A, Teichmann SA.
-\newblock Putative cell type discovery from single-cell gene expression data.
-\newblock Nature Methods. 2020;17(6):621--628.
-\newblock doi:{10.1038/s41592-020-0825-9}.
-
-\bibitem{holland_1983}
-Holland PW, Laskey KB, Leinhardt S.
-\newblock Stochastic blockmodels: First steps.
-\newblock Social networks. 1983;5(2):109--137.
-\newblock doi:{10.1016/0378-8733(83)90021-7}.
-
-\bibitem{peixoto_2017}
-Peixoto TP.
-\newblock Nonparametric Bayesian inference of the microcanonical stochastic
-  block model.
-\newblock Physical review E. 2017;95(1-1):012317.
-\newblock doi:{10.1103/{PhysRevE}.95.012317}.
-
-\bibitem{karrer_2011}
-Karrer B, Newman MEJ.
-\newblock Stochastic blockmodels and community structure in networks.
-\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
-  2011;83(1 Pt 2):016107.
-\newblock doi:{10.1103/{PhysRevE}.83.016107}.
-
-\bibitem{peixoto_2013}
-Peixoto TP.
-\newblock Parsimonious module inference in large networks.
-\newblock Physical Review Letters. 2013;110(14):148701.
-\newblock doi:{10.1103/{PhysRevLett}.110.148701}.
-
-\bibitem{peixoto_2014}
-Peixoto TP.
-\newblock Efficient Monte Carlo and greedy heuristic for the inference of
-  stochastic block models.
-\newblock Physical Review E, Statistical, Nonlinear, and Soft Matter Physics.
-  2014;89(1):012804.
-\newblock doi:{10.1103/{PhysRevE}.89.012804}.
-
-\bibitem{peixoto_2014_h}
-Peixoto TP.
-\newblock Hierarchical Block Structures and High-Resolution Model Selection in
-  Large Networks.
-\newblock Physical Review X. 2014;4(1):011047.
-\newblock doi:{10.1103/{PhysRevX}.4.011047}.
-
-\bibitem{tian_2019}
-Tian L, Dong X, Freytag S, Lê~Cao KA, Su S, {JalalAbadi} A, et~al.
-\newblock Benchmarking single cell {RNA}-sequencing analysis pipelines using
-  mixture control experiments.
-\newblock Nature Methods. 2019;16(6):479--487.
-\newblock doi:{10.1038/s41592-019-0425-8}.
-
-\bibitem{kang_2018}
-Kang HM, Subramaniam M, Targ S, Nguyen M, Maliskova L, {McCarthy} E, et~al.
-\newblock Multiplexed droplet single-cell {RNA}-sequencing using natural
-  genetic variation.
-\newblock Nature Biotechnology. 2018;36(1):89--94.
-\newblock doi:{10.1038/nbt.4042}.
-
-\bibitem{mereu_2020}
-Mereu E, Lafzi A, Moutinho C, Ziegenhain C, {McCarthy} DJ, {Álvarez-Varela} A,
-  et~al.
-\newblock Benchmarking single-cell {RNA}-sequencing protocols for cell atlas
-  projects.
-\newblock Nature Biotechnology. 2020;38(6):747--755.
-\newblock doi:{10.1038/s41587-020-0469-4}.
-
-\bibitem{sasagawa_2018}
-Sasagawa Y, Danno H, Takada H, Ebisawa M, Tanaka K, Hayashi T, et~al.
-\newblock Quartz-Seq2: a high-throughput single-cell {RNA}-sequencing method
-  that effectively uses limited sequence reads.
-\newblock Genome Biology. 2018;19(1):29.
-\newblock doi:{10.1186/s13059-018-1407-3}.
-
-\bibitem{zheng_2017}
-Zheng GXY, Terry JM, Belgrader P, Ryvkin P, Bent ZW, Wilson R, et~al.
-\newblock Massively parallel digital transcriptional profiling of single cells.
-\newblock Nature Communications. 2017;8:14049.
-\newblock doi:{10.1038/ncomms14049}.
-
-\bibitem{jaitin_2014}
-Jaitin DA, Kenigsberg E, Keren-Shaul H, Elefant N, Paul F, Zaretsky I, et~al.
-\newblock Massively parallel single-cell {RNA}-seq for marker-free
-  decomposition of tissues into cell types.
-\newblock Science. 2014;343(6172):776--779.
-\newblock doi:{10.1126/science.1247651}.
-
-\bibitem{goldstein_2017}
-Goldstein LD, Chen YJJ, Dunne J, Mir A, Hubschle H, Guillory J, et~al.
-\newblock Massively parallel nanowell-based single-cell gene expression
-  profiling.
-\newblock {BMC} Genomics. 2017;18(1):519.
-\newblock doi:{10.1186/s12864-017-3893-1}.
-
-\bibitem{paul_2015}
-Paul F, Arkin Y, Giladi A, Jaitin DA, Kenigsberg E, Keren-Shaul H, et~al.
-\newblock Transcriptional heterogeneity and lineage commitment in myeloid
-  progenitors.
-\newblock Cell. 2015;163(7):1663--1677.
-\newblock doi:{10.1016/j.cell.2015.11.013}.
-
-\bibitem{wolf_2019}
-Wolf FA, Hamey FK, Plass M, Solana J, Dahlin JS, {Göttgens} B, et~al.
-\newblock {PAGA}: graph abstraction reconciles clustering with trajectory
-  inference through a topology preserving map of single cells.
-\newblock Genome Biology. 2019;20(1):59.
-\newblock doi:{10.1186/s13059-019-1663-x}.
-
-\bibitem{peixoto_2020}
-Peixoto TP.
-\newblock Merge-split Markov chain Monte Carlo for community detection.
-\newblock arXiv. 2020;doi:{arXiv:2003.07070}.
-
-\end{thebibliography}
+\bibliography{references}
 

### 2020-11-16 supplementary.tex
initial import of old files

--- 2020-11-16 GenomeBiology/supplementary.tex
@@
 }
-\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis).
-Both the parameters influence the sparseness of the graph.}
+Both the parameters influence the sparseness of the graph.}\label{MPFigure:F56BBDC3-0F0A-4558-8BC2-839E91250DB4}
 \end{figure}
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
-\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}
+\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run.
+Maximal ARI over all hierarchy level is shown.
+Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
 \end{figure}
@@
 (A) Low dimensional embedding of single cells colored by original cell type and pseudotime.
-(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}
+(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{MPFigure:6C191331-D4E9-4774-A8E0-210B2852D498}
 \end{figure}
@@
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution $r=0.3$ splits in groups 2 and 0 at resolution $r=0.2$.}
+Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution \emph{r}=0.3 splits in groups 2 and 0 at resolution \emph{r}=0.2.}\label{MPFigure:76168960-F76F-44B1-A5FA-A31681CDAFFD}
 \end{figure}
@@
 \caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency.
-Dot size is proportional to the number of cells remaining after filtering.}
+Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
 \end{figure}

### 2021-04-14 manuscript.tex
New sections in place

@@
 \begin{document}
-\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
+\title{Stochastic Block Models Applied to the Analysis of Single Cell Data}
 \author[1,2]{Leonardo Morelli}
@@
 Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity.
-In this paper we propose an alternative solution to this problem, based on nested Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.
+In this paper we propose an alternative solution to this problem, based on Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.
 
@@
 
-We developed \emph{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \emph{scanpy}, to facilitate the adoption of nested stochastic block models in single-cell analysis.
+We developed \emph{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \emph{scanpy}, to facilitate the adoption of stochastic block models in single-cell analysis.
 
@@
 \subsection*{nSBM does not return artificial cell groupings}
+
+\textcolor{red}{TO BE UPDATED}
+
 One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling.
@@
 \subsection*{nSBM correctly identifies cell populations}
+
+\textcolor{red}{TO BE UPDATED, possibly with some simulation study}
+
 To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform.
@@
 
-\subsection*{Cell affinities can be used to evaluate cluster purity}
-The computational framework underlying \emph{schist} calculates the model entropy, that is the amount of information required to describe a block configuration.
-Given that minimisation of such quantity can be used to perform model selection, it can be also used to evaluate the impact of modifying the assignment of a cell to a cluster.
-Once a model is minimised, \emph{schist} performs an exhaustive exploration of all model entropies resulting from moving all cells into all possible clusters.
-The differences in entropies could be interpreted as affinities of cells to given clusters.
-Such affinities are, in fact, probability values and could be used to evaluate the internal consistency of a given cell cluster.
+\subsection*{Comparison with other methods}
 
-\begin{figure}[H]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
-\caption[]{Analysis of cell cluster consistency.
-Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform.
-Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1.
-The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster.
-Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
-\end{figure}
-
-To this end we calculate the entropy of the group-wise distribution of cell affinities, which is maximal when all cells have affinity equal to 1 for a given group.
-We tested this idea on four datasets recently published to benchmark single cell technologies in the Human Cell Atlas project \cite{mereu_2020}; in particular, we chose two technologies resulting in high quality data: Quartz-seq2 \cite{sasagawa_2018} and Chromium 10x v3 \cite{zheng_2017}, and two technologies resulting in more noisy data: MARS-seq \cite{jaitin_2014} and iCell8 \cite{goldstein_2017} (Fig.
-\ref{Figure3}).
-
-Cluster consistency is not a measure of the data quality, in fact we identify low consistency groups in all datasets.
-High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper.
-Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells (Fig.
-S5).
-
-Similarly, we can use cell affinities to derive a stability parameter, a measure of the tendency for a cell to be stably associated to given clusters at all levels of the hierarchy.
-To this end, we first calculate the cell-wise entropy $H_{i,h}$ of cell affinity at each hierarchy level \emph{h}, then we define the stability as $S_i = 1 - \max(H_i)$.
-While we conceived this measure to identify and exclude cells with dubious assignment, we found that it may be more useful to assess the general data quality: the fraction of cells having $S>0.95$ was 0.783, 0.795, 0.831 and 0.855 for the iCELL8, MARS-seq, Chromium 10x and Quartz-seq2 technology respectively, in line with the evaluation on increasing performances of those platforms in \cite{mereu_2020}.
+\textcolor{red}{TO BE FILLED with SSCAF?}
 
@@
 We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.
+\subsection*{Cell affinities can be used for label transfer}
+
+\textcolor{red}{TO BE DONE}
+
+\subsection*{PPBM?}
+
+\textcolor{red}{TO BE DONE}
 
 \subsection*{Analysis of runtimes}
+
+\textcolor{red}{TO BE UPDATED WITH NEW DATA}
+
 Minimisation of the nSBM is a process that may require a large amount of computational resources.

### 2021-04-15 manuscript.tex
Some ideas about paper narrative

@@
 \subsection*{Model hierarchy contains biological information}
+
+\textcolor{red}{TO BE UPDATED, possibly with Visium data. The point here should be that hierarchy is more consistent with tissue type, then we talk about development}
 The hierarchical model of cell groups implies that a relationship exists between groups.
@@
 
-\textcolor{red}{TO BE DONE}
+\textcolor{red}{TO BE DONE. Maybe this is the right place to introduce the choice of the appropriate hierarchy level}
 
@@
 \centering
- \begin{tabular}{|| l c c c c  ||}
+ \begin{tabular}{|| l c c c c c ||}
 %  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
  \hline
- \textbf{Dataset} & \textbf{Cells} & \textbf{Minimize} & \textbf{Merge-split MCMC} & \textbf{Overlap} \\ [0.5ex] 
+ \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{NSBM} \\ [0.5ex] 
  \hline\hline
- sc-mixology \cite{tian_2019} & 860 & 01:11 & 00:03 & 0.884 \\ 
+% sc-mixology \cite{tian_2019} & 860 & 01:11 & 00:03 & 0.884 \\ 
+ %\hline
+ Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 
  \hline
- Quartzseq \cite{mereu_2020} & 1266 & 00:31 & 00:02 & 0.726 \\ 
+ Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:16 & 00:30 & 02:10 \\
  \hline
- MARS-seq \cite{mereu_2020} & 1401 & 00:29 & 00:08 & 0.834 \\
+ MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:24 & 01:10 & 06:11 \\
  \hline
- Chromium 10X \cite{mereu_2020} & 1523 & 00:40 & 00:04 & 0.695 \\
+ iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:34 & 01:31 & 09:06 \\
  \hline
- iCELL8 \cite{mereu_2020} & 1830 & 00:54 & 00:07 & 0.623 \\
+ Paul15 \cite{paul_2015} & 2730 & 51475 & 01:21 & 03:35 & 12:17\\ 
  \hline
- Paul15 \cite{paul_2015} & 2730 & 02:37 & 00:10 & 0.575 \\ 
- \hline
- Planaria \cite{plass_2018} & 21612 & 13:12 & 03:29 & 0.589 \\
+ Planaria \cite{plass_2018} & 21612 & 173667 & 06:32 & 34:12 & 3:17:08 \\
  \hline

### 2021-04-15 manuscript.tex
updated introduction and random data

@@
 \begin{document}
-\title{Stochastic Block Models Applied to the Analysis of Single Cell Data}
+\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
 \author[1,2]{Leonardo Morelli}
@@
 
-A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, groups of cells sharing similar properties.
-Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}.
+A key step in the analysis of scRNA-seq data and, more in general, of single cell data, is the identification of cell populations, that is groups of cells sharing similar properties.
+Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017, Ranjan_2021} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}.
 As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards.
-Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN) which is then partitioned into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
+Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN or shared Nearest Neighbors, \emph{s}NN).
+Encoding cell-to-cell similarities into graphs has practical advantages, as many algorithms for graph analysis can be applied and interpreted in a biological way.
+A notable example is the analysis of cell trajectories which can be derived from the analysis of random walks on the NN graph (\textcolor{red}{CHECK THIS: it's actually from diffusion maps}; computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}.
+Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
 While fast, this method does not guarantee that small communities in large networks are well defined.
-To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy and PhenoGraph \cite{levine_2015}.
-In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006} .
+To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy \cite{wolf_2018} and PhenoGraph \cite{levine_2015}.
+In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006}.
 In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}.
@@
 Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model.
-Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019} or clique analysis \cite{xu_2015}.
+Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}.
 Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.
@@
 \end{MPEquation}
+
 so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
@@
 In addition, it is possible to measure the confidence of assignment to groups.
-We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.
+\textcolor{red}{We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.}
 
@@
 When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph.
-\emph{schist} annotates cells in the data object with all the groups found at each level of a hierarchy.
-As there could be more model fits with similar entropy, \emph{schist} could explore the space of solutions with a Markov Chain Monte Carlo (MCMC) algorithm, to perform model averaging; this step is performed until it converges, that is the difference in model entropy in \emph{n} continuous iterations remains under a specified threshold.
-Sampling from the posterior distribution can be used to study the distribution of the number of groups, at each level.
-This information (group marginals) could be studied to identify the most probable number of cell groups.
-Once \emph{schist} has fitted a model, it evaluates the difference in entropy given by assigning every cell to every possible group.
-This step generates the matrix of \emph{cell affinity}, that is the probability for a cell to belong to a specific group.
-A cell affinity matrix is generated and returned for every hierarchy level.
-We show that this information can be useful to evaluate cluster consistence.
-
-%As the core object of \emph{graph-tool }used by \emph{schist}, the \emph{NestedBlockState}, cannot be serialised by \emph{scanpy}'s IO infrastructure, we implemented simple read\slash write functions to dump \emph{AnnData} is h5ad format, together with the \emph{NestedBlockState} as a separate file using python's \emph{pickle} library.
+\emph{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy.
+Given the large size of the NN graph, it is possible that a single solution represents local minima of the fitting process.
+In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}.
+To overcome these issues, \emph{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}).
+In addition to the default Nested Stochastic Block Model, \emph{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
+Once a model has been fitted, \emph{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
+This step allows to estimate the probability for a cell to belong to a specific group (\emph{cell affinity}).
+We show that cell affinity may have practical application in common tasks in single cell analysis, such as label transfer.
 
 \subsection*{nSBM does not return artificial cell groupings}
-
-\textcolor{red}{TO BE UPDATED}
 
@@
 To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged.
-We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0431$, $p=0.993$).
+We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$).
 We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.
@@
 (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right).
-Cell grouping is consistent for both the approaches ($ARI=0.804$).
+Cell grouping is consistent for both the approaches (Adjusted Rand Index $ARI=0.869$).
 (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges.

### 2021-04-15 manuscript.tex
Added Visium data

@@
 
-\subsection*{Model hierarchy contains biological information}
+\subsection*{Hierarchy modelling is accurate in defining biological properties}
+
+When grouping is performed by optimization of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions.
+Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
+To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package SquidPy \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors.
+At default resolution, modularity optimization resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig.
+\ref{Figure_Visium}).
+When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.
+
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Visium_Spatial.png}
+\caption[]{Analysis of spatial transcriptomics of cerebellum by Visium HD.
+In the first panel, original tissue annotation is given.
+Tissues are well defined at default resolution for the standard approach.
+When resolution is decreased to $\gamma = 0.5$, cells from the dentate gyrus are merged to the hippocampus.
+When nSBM is used, the structure of the pyramidal is maintained at different levels of the hierarchy}\label{Figure_Visium}
+\end{figure}
 

### 2021-04-15 manuscript.tex
initial rewrite of developmental

@@
 \usepackage[numbers]{natbib}
+\usepackage{xr}
+\externaldocument{supplementary}
+
 \bibliographystyle{plos2015}
@@
 
-To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform.
+%To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform.
 At a first evaluation of the UMAP embedding, all lines appear well separated.
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
-\caption[]{\emph{schist} applied to scRNA-seq mixology data.
+%\begin{figure}[H]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
+%\caption[]{\emph{schist} applied to scRNA-seq mixology data.
 (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper.
@@
 The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
-\end{figure}
+%\end{figure}
 
-Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig.
+%Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig.
 \ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$).
@@
 It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
-As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig.
+%As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig.
 \ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.
 
-These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter.
+%These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter.
 These data also hint at the possibility to identify rare cell types in larger populations.
@@
 \ref{Figure_Visium}).
-When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.
+When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer, suggesting that the standard approach may introduce biases.
 
@@
 
-\textcolor{red}{TO BE UPDATED, possibly with Visium data. The point here should be that hierarchy is more consistent with tissue type, then we talk about development}
-The hierarchical model of cell groups implies that a relationship exists between groups.
-We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation.
-To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}.
-Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig.
-S3A).
-After applying nSBM, we identify 27 groups at level 3 of the hierarchy (Fig.
-S3B), compared to the 24 using Leiden method at default resolution (Fig.
-S4).
-We found that the hierarchy proposed by our model is consistent with the developmental model (Fig.
-\ref{Figure2}).
-Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (Fig.
-S4); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.
+In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameters.
+We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}.
+Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig.
+\textcolor{red}{S3A}).
+After applying Leiden method at default resolution we identified 24 groups.
+By lowering the $\gamma$ parameter we could observe cell groups that merge and split at different resolutions in a non hierarchcal manner (Fig.
+\ref{Figure_Paul15_Leiden_r}).
 

### 2021-04-15 supplementary.tex
initial rewrite of developmental

@@
 }
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S4.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Paul15_Leiden_r.png}
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ }group 3 at resolution \emph{r}=0.3 splits in groups 2 and 0 at resolution \emph{r}=0.2.}\label{MPFigure:76168960-F76F-44B1-A5FA-A31681CDAFFD}
+Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ } in the Erythroid branch groups 7 and 15 at resolution \emph{r}=1 are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
 \end{figure}

### 2021-04-15 manuscript.tex
label transfer

@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
-\caption[]{Analysis of hematopoietic differentiation.
+%\begin{figure}[H]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
+%\caption[]{Analysis of hematopoietic differentiation.
 Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy.
@@
 In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
-\end{figure}
+%\end{figure}
 
-In all, these data suggest that not only nSBM is able to identify consistent cell groups at different scales, but also that the hierarchy proposed by the model has a direct biological interpretation.
+In all, these data suggest that \textcolor{red}{SAY SOMETHING HERE}.
 
@@
 
-\textcolor{red}{TO BE DONE}
+The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme.
+Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data.
+We sought to exploit this property to address the task of annotating cells according to a reference sample.
+To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies.
+We choose cells profiled with Chromium 10X and MARS-seq platforms ((Fig.
+\ref{Figure_Label_Transfer}A-B), as those were found to be at the extremes of capability to distinguish cell types.
+After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated both datasets into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph from the resulting unified representation.
+Note that the original cell type is known for both datasets.
+We retained cell type annotation for cells derived from 10X, while we assigned a 'Unknown' label to all cells derived from MARS-seq.
+We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'.
+Once the matrix has been computed, each cell from the MARS-seq data is assigned to the group with the highest likelihood.
+The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they are annotated as different cell types; cells from the MARS-seq should retain their unannotated label if and only if there is not enough evidence to associate them to another group.
+In parallel, we also relabeled MARS-seq data according to the closest annotated cell in the reference \emph{k}NN graph.
+Finally we calculated the accuracy of label transfer of both methods using known annotations; we found that the \emph{k}NN approach (Fig.
+\ref{Figure_Label_Transfer}C) performs worse than an approach based on Stochastic Block Models (Fig.
+\ref{Figure_Label_Transfer}C).
+In the latter, approximately 4\% of cells remain unlabeled.
+Interestingly, we found that for the largest part of these cells, the second choice by affinity was indeed the appropriate one (Fig.
+\ref{hist_label_second}).
+
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Label_Transfer.png}
+\caption[]{Label transfer using SBM.
+Annotated data from Chromium 10X (A) and  MARS-seq (B) were merged using Harmony.
+MARS-seq cells were relabeled according to the \emph{k}NN graph (C) or by exploiting the Stochastic Block Model (D).
+}\label{Figure_Label_Transfer}
+\end{figure}
 

### 2021-04-15 supplementary.tex
label transfer

@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
-\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors.
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
+%\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors.
 Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis).
 Both the parameters influence the sparseness of the graph.}\label{MPFigure:F56BBDC3-0F0A-4558-8BC2-839E91250DB4}
-\end{figure}
+%\end{figure}
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
-\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run.
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
+%\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run.
 Maximal ARI over all hierarchy level is shown.
 Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
-\end{figure}
+%\end{figure}
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
-\caption[]{Analysis of hematopoietic differentiation.
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
+%\caption[]{Analysis of hematopoietic differentiation.
 (A) Low dimensional embedding of single cells colored by original cell type and pseudotime.
 (B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{MPFigure:6C191331-D4E9-4774-A8E0-210B2852D498}
-\end{figure}
+%\end{figure}
 
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
-\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency.
-Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{hist_label_second.png}
+\caption[]{TO BE WRITTEN}\label{hist_label_second}
 \end{figure}
 
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
+%\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency.
+Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
+%\end{figure}
+
 \end{document}

### 2021-04-15 manuscript.tex
new section
Hierarchies part 2

@@
 In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameters.
-We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}.
-Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig.
-\textcolor{red}{S3A}).
+We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019} (Fig.
+\ref{Figure_Hemato_Supp}A).
+Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells, mostly recapitulated by level 2 of the hierarchy computed by \emph{schist} (Fig.
+\ref{Figure_Hemato_Lineage}).
+Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime.
 After applying Leiden method at default resolution we identified 24 groups.
-By lowering the $\gamma$ parameter we could observe cell groups that merge and split at different resolutions in a non hierarchcal manner (Fig.
+By lowering the $\gamma$ parameter we could observe cell groups that merge and split at different resolutions disrupting the hierarchy (Fig.
 \ref{Figure_Paul15_Leiden_r}).
 
-%\begin{figure}[H]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
-%\caption[]{Analysis of hematopoietic differentiation.
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Hemato_Lineage.png}
+\caption[]{Analysis of hematopoietic differentiation.
 Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy.
-Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A).
-In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
-%\end{figure}
+Cells are colored according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A).
+In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure_Hemato_Lineage}
+\end{figure}
 
-In all, these data suggest that \textcolor{red}{SAY SOMETHING HERE}.
+In all, these data show that the common intuition that $\gamma$ parameter acts as a thresholding factor over a hierarchy is wrong.
+Not only the hierarchy is not conserved, but also very different cell types may be mixed in spurious clusters.
+Conversely, the nSBM is able to represent hierarchical relations in appropriate way.
+Moreover, the hierarchy appears to be more robust in aggregating different cell types at coarser scales.
 
@@
 \end{figure}
+
+\subsection*{Choice of the optimal level}
+
+\textcolor{red}{TO BE DONE. Discuss modularity, but also eigenanalysis of random matrix}.
 
@@
 \section*{Conclusions}
+\textcolor{red}{TO BE UPDATED}
 

### 2021-04-15 supplementary.tex
Hierarchies part 2

@@
 
-%\begin{figure}[htbp]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S3.png}
-%\caption[]{Analysis of hematopoietic differentiation.
+\begin{figure}[htbp]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.png}
+\caption[]{Analysis of hematopoietic differentiation.
 (A) Low dimensional embedding of single cells colored by original cell type and pseudotime.
-(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{MPFigure:6C191331-D4E9-4774-A8E0-210B2852D498}
-%\end{figure}
+(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
+\end{figure}
 
@@
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ } in the Erythroid branch groups 7 and 15 at resolution \emph{r}=1 are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
+Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ } in the Erythroid branch groups \textcolor{red}{XXX and XXX} at resolution \emph{r}=1 are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
 \end{figure}

### 2021-04-16 manuscript.tex
udpate runtimes for paul15
New figure for label transfer

@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Hemato_Lineage.png}
+\includegraphics[keepaspectratio,width=\textwidth,width=0.55\textheight]{Figure_Hemato_Lineage.png}
 \caption[]{Analysis of hematopoietic differentiation.
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Label_Transfer.png}
+\includegraphics[keepaspectratio,width=\textwidth,width=0.55\textheight]{Figure_Label_Transfer.png}
 \caption[]{Label transfer using SBM.
@@
  \hline
- Paul15 \cite{paul_2015} & 2730 & 51475 & 01:21 & 03:35 & 12:17\\ 
+ Paul15 \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
  \hline

### 2021-04-16 manuscript.tex
some mods

@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,width=0.55\textheight]{Figure_Hemato_Lineage.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.7\textheight]{Figure_Hemato_Lineage.png}
 \caption[]{Analysis of hematopoietic differentiation.
@@
 To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies.
-We choose cells profiled with Chromium 10X and MARS-seq platforms ((Fig.
-\ref{Figure_Label_Transfer}A-B), as those were found to be at the extremes of capability to distinguish cell types.
-After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated both datasets into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph from the resulting unified representation.
-Note that the original cell type is known for both datasets.
-We retained cell type annotation for cells derived from 10X, while we assigned a 'Unknown' label to all cells derived from MARS-seq.
+We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 and MARS-seq.
+These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.
+After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph.
+In each merged dataset, we retained cell type annotations for 10XV3 cells, while we assigned a 'Unknown' label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2).
 We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'.
-Once the matrix has been computed, each cell from the MARS-seq data is assigned to the group with the highest likelihood.
-The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they are annotated as different cell types; cells from the MARS-seq should retain their unannotated label if and only if there is not enough evidence to associate them to another group.
-In parallel, we also relabeled MARS-seq data according to the closest annotated cell in the reference \emph{k}NN graph.
-Finally we calculated the accuracy of label transfer of both methods using known annotations; we found that the \emph{k}NN approach (Fig.
-\ref{Figure_Label_Transfer}C) performs worse than an approach based on Stochastic Block Models (Fig.
-\ref{Figure_Label_Transfer}C).
-In the latter, approximately 4\% of cells remain unlabeled.
-Interestingly, we found that for the largest part of these cells, the second choice by affinity was indeed the appropriate one (Fig.
+Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood.
+The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their unannotated label if and only if there is not enough evidence to associate them to another group.
+We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \emph{ingest}, a tool included in Scanpy also based on \emph{k}NN classification.
+Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig.
+\ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927.
+When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unknown.
+This effect is particular evident for \emph{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243).
+Conversely, \emph{schist} is able to assign correct labels with higher accuracy (0.641).
+Moreover, \emph{k}NN methods assign a label to each cell, whereas \emph{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the Unknown state is the most likely).
+Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity was indeed the appropriate one (Fig.
 \ref{hist_label_second}).
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,width=0.55\textheight]{Figure_Label_Transfer.png}
+\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_Label_Transfer.png}
 \caption[]{Label transfer using SBM.

### 2021-04-16 supplementary.tex
some mods

@@
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.\ } in the Erythroid branch groups \textcolor{red}{XXX and XXX} at resolution \emph{r}=1 are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
+Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.} groups 6 and 8 in the Erythroid branch at resolution $\gamma=1$ are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
 \end{figure}

### 2021-04-19 manuscript.tex
mods
changed font layout

@@
 Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system.
-While properties of single cells is the primary endpoint of such analysis, these are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation.
+The definition of the  properties of single cells is the primary endpoint of such analysis, nevertheless cells are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation.
 Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity.
-In this paper we propose an alternative solution to this problem, based on Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.
+In this paper we propose an alternative and principled solution to this problem, based on Stochastic Block Models.
+We show that our approach is not only able to correctly identify cell groups, but also adds the possibility to perform inference on cell identities \textcolor{red}{SOMETHING MORE?
+Damn, I always struggle with abstracts}.
 
@@
 As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards.
-Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN or shared Nearest Neighbors, \emph{s}NN).
-Encoding cell-to-cell similarities into graphs has practical advantages, as many algorithms for graph analysis can be applied and interpreted in a biological way.
-A notable example is the analysis of cell trajectories which can be derived from the analysis of random walks on the NN graph (\textcolor{red}{CHECK THIS: it's actually from diffusion maps}; computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}.
+Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbors, \emph{s}NN).
+Encoding cell-to-cell similarities into graphs has practical advantages beyond clustering, as many algorithms for graph analysis can be applied and interpreted in a biological way.
+A notable example is the analysis of cell trajectories which can be derived from the analysis of Markov processes on the NN graph \cite{Setty_2019, Lange_2020} ; computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}.
 Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
@@
 Moreover, it has been demonstrated that random networks can have modularity \cite{guimer_2004} and its optimisation is incapable of separating actual structure from those arising simply of statistical fluctuations of the null model.
+Lastly, it is a common error to assume that the resolution parameter reflects a hierarchical structure of the communities in the graph when, in general, this is not rigorously true.
 Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}.
-Lastly, it has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.
+It has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.
 
 An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organized into communities \cite{holland_1983}.
-In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed neighbourhood graph.
+In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed NN graph.
 According to the microcanonical formulation \cite{peixoto_2017}, the parameters are node partitions into groups and the matrix of edge counts between groups themselves.
-Under this model, nodes belonging to the same group have the same probability to be connected with other nodes.
+Under this model, nodes belonging to the same group have the same probability to be connected together.
 It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs.
@@
 \end{equation}
-\label{MPEquationElement:820CB015-E51E-4C5A-BD9C-70E667B8F73B}
+\label{Equation_posterior}
 \end{MPEquation}
@@
 \end{equation}
-\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
+\label{Equation_entropy}
 \end{MPEquation}
 
-so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
+so that inference is performed by minimizing the information required to describe the data (Occam's razor); \texttt{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
 SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}.
@@
 
-We developed \emph{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \emph{scanpy}, to facilitate the adoption of stochastic block models in single-cell analysis.
+We developed \texttt{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \texttt{scanpy}, to facilitate the adoption of stochastic block models in single-cell analysis.
 
@@
 
-\subsection*{Overview of \emph{schist}}
+\subsection*{Overview of \texttt{schist}}
 
-\emph{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \emph{scanpy}.
-The most prominent function is \emph{schist.inference.nested\_model()} which takes a \emph{AnnData} object as input and fits a nested stochastic block model on the \emph{k}NN graph built with \emph{scanpy} functions (\emph{e.g.\ scanpy.tools.neighbors()}).
-When launched with default parameters, \emph{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph.
-\emph{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy.
+\texttt{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}.
+The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}).
+When launched with default parameters, \texttt{schist} fits a model which maximises the posterior probability of having a set of cell groups (or blocks) given a graph.
+\texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy.
 Given the large size of the NN graph, it is possible that a single solution represents local minima of the fitting process.
 In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}.
-To overcome these issues, \emph{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}).
-In addition to the default Nested Stochastic Block Model, \emph{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
-Once a model has been fitted, \emph{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
-This step allows to estimate the probability for a cell to belong to a specific group (\emph{cell affinity}).
+To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}).
+In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
+Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
+This step allows to estimate the strength of association of a cell to a specific group (\emph{cell affinity}).
 We show that cell affinity may have practical application in common tasks in single cell analysis, such as label transfer.
 
-\subsection*{nSBM does not return artificial cell groupings}
+\subsection*{\texttt{schist} does not return artificial cell groupings}
 
-One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling.
+One of the most relevant difference between SBM and other methods to cluster single cells is that it relies on robust statistical modelling.
 In this sense, the number of groups identified strictly mirrors the amount of information contained in the data.
@@
 To show this property we performed a simple experiment on a randomised \emph{k}NN graph.
-To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged.
+We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged.
 We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$).
@@
 Nevertheless, real-world data may include an unknown amount of random noise.
-Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.
+Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
 
-\subsection*{nSBM correctly identifies cell populations}
+\subsection*{\texttt{schist} correctly identifies cell populations}
 
@@
 Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
-To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package SquidPy \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors.
+To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package \texttt{SquidPy} \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors.
 At default resolution, modularity optimization resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig.
 \ref{Figure_Visium}).
-When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer, suggesting that the standard approach may introduce biases.
+When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.
 
@@
 \ref{Figure_Hemato_Supp}A).
-Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells, mostly recapitulated by level 2 of the hierarchy computed by \emph{schist} (Fig.
+Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells, mostly recapitulated by level 2 of the hierarchy computed by \texttt{schist} (Fig.
 \ref{Figure_Hemato_Lineage}).
 Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime.
-After applying Leiden method at default resolution we identified 24 groups.
-By lowering the $\gamma$ parameter we could observe cell groups that merge and split at different resolutions disrupting the hierarchy (Fig.
+Conversely, the Leiden method at default resolution identified 24 groups.
+By lowering the $\gamma$ parameter we observed cell groups that merge and split at different resolutions disrupting the hierarchy (Fig.
 \ref{Figure_Paul15_Leiden_r}).
@@
 Not only the hierarchy is not conserved, but also very different cell types may be mixed in spurious clusters.
-Conversely, the nSBM is able to represent hierarchical relations in appropriate way.
+By using SBM, \texttt{schist} is able to represent hierarchical relations in appropriate way.
 Moreover, the hierarchy appears to be more robust in aggregating different cell types at coarser scales.
@@
 \subsection*{Analysis of runtimes}
-
-\textcolor{red}{TO BE UPDATED WITH NEW DATA}
 

### 2021-04-19 manuscript.tex
added a section for cell marginals

@@
 We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.
+
+\subsection*{Cell marginals improve detection of marker genes}
+
+\textcolor{red}{TO BE DONE.}.
+
 \subsection*{Cell affinities can be used for label transfer}
@@
 The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their unannotated label if and only if there is not enough evidence to associate them to another group.
-We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \emph{ingest}, a tool included in Scanpy also based on \emph{k}NN classification.
+We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in Scanpy also based on \emph{k}NN classification.
 Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig.
 \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927.
-When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unknown.
-This effect is particular evident for \emph{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243).
-Conversely, \emph{schist} is able to assign correct labels with higher accuracy (0.641).
-Moreover, \emph{k}NN methods assign a label to each cell, whereas \emph{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the Unknown state is the most likely).
+When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabeled cells.
+This effect is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243).
+Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641).
+Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely).
 Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity was indeed the appropriate one (Fig.

### 2021-04-20 manuscript.tex
some changes

@@
 
-%To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform.
+To benchmark \texttt{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data.
+In particular, we used a mixture of 5 cell lines profiled with Chromium 10x platform.
 At a first evaluation of the UMAP embedding, all lines appear well separated.
-Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig.
-\ref{Figure1}A).
+Only the lung cancer line H1975 shows a considerable degree of heterogeneity and appears to be split into two cell groups (Fig.
+\ref{FigureTian}A).
+Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.854$), with a further split in H2228 cell line (Fig.
+\ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.565$).
+
  Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph.
@@
 
-%\begin{figure}[H]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
-%\caption[]{\emph{schist} applied to scRNA-seq mixology data.
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,width=0.5\textheight]{FigureTian.png}
+\caption[]{\emph{schist} applied to scRNA-seq mixology data.
 (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper.
@@
 (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below).
-The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
-%\end{figure}
+The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{FigureTian}
+\end{figure}
 
@@
 
+\textcolor{red}{TO BE UPDATED}
+
 \subsection*{Analysis of Random data}

### 2021-04-20 manuscript.tex
figure missing
use marginals for gene xp
Tian done

@@
 Several approaches have been proposed to achieve this task, based on well established clustering techniques \cite{wang_2017, lin_2017}, consensus clustering \cite{huh_2020, kiselev_2017, Ranjan_2021} and deep learning \cite{li_2020}; many more have been recently reviewed \cite{krzak_2019, kiselev_2019} and benchmarked \cite{du_2018}.
-As the popularity of single cell analysis frameworks Seurat \cite{butler_2018} and Scanpy \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards.
+As the popularity of single cell analysis frameworks \texttt{Seurat} \cite{butler_2018} and \texttt{scanpy} \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards.
 Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbors, \emph{s}NN).
 Encoding cell-to-cell similarities into graphs has practical advantages beyond clustering, as many algorithms for graph analysis can be applied and interpreted in a biological way.
-A notable example is the analysis of cell trajectories which can be derived from the analysis of Markov processes on the NN graph \cite{Setty_2019, Lange_2020} ; computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}.
-Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; the latter step is typically performed using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
+A notable example is the analysis of cell trajectories which can be derived from the analysis of Markov processes traversing the NN graph \cite{Setty_2019, Lange_2020}.
+In another context, computation of RNA moments in scRNA velocity is also based on the NN graph structure \cite{Bergen_2020}.
+Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; this is typically achieved using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
 While fast, this method does not guarantee that small communities in large networks are well defined.
-To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by Scanpy \cite{wolf_2018} and PhenoGraph \cite{levine_2015}.
+To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by \texttt{scanpy} \cite{wolf_2018} and \texttt{PhenoGraph} \cite{levine_2015}.
 In addition to Newman's modularity \cite{newman_2004}, other definitions currently used in single cell analysis make use of a resolution parameter \cite{traag_2011, reichardt_2006}.
-In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{viceversa}.
+In lay terms, resolution works as a threshold on the density within communities: lowering the resolution results in less and sparser communities and \emph{vice versa}.
 Identification of an appropriate resolution has been recognised as a major issue \cite{lhnemann_2020}, also because it requires the definition of a mathematical property (clusters) over biological entities (the cell groups), with little formal description of the latter.
@@
 Lastly, it is a common error to assume that the resolution parameter reflects a hierarchical structure of the communities in the graph when, in general, this is not rigorously true.
-Additional solutions to cell group identification from neighbourhood graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}.
+Additional solutions to cell group identification from NN graphs have been proposed, introducing resampling techniques \cite{baran_2019, Tang_Sackton_2020} or clique analysis \cite{xu_2015}.
 It has been proposed that high resolution clustering, \emph{e.g.} obtained with Leiden or Louvain methods, can be refined in agglomerative way using machine learning techniques \cite{miao_2020}.
@@
 In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed NN graph.
-According to the microcanonical formulation \cite{peixoto_2017}, the parameters are node partitions into groups and the matrix of edge counts between groups themselves.
+According to the microcanonical formulation \cite{peixoto_2017}, the parameters are partitions and the matrix of edge counts between them.
 Under this model, nodes belonging to the same group have the same probability to be connected together.
 It is possible to include node degree among the model parameters \cite{karrer_2011}, to account for heterogeneity of degree distribution of real-world graphs.
-A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \emph{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}).
+A Bayesian approach to infer parameters has been developed \cite{peixoto_2013} and implemented in the \texttt{graph-tool} python library (\href{https://graph-tool.skewed.de}{https:/\slash graph-tool.skewed.de}).
 There, a generative model of network $\boldsymbol A$ has a probability $P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)$ where \textbf{$\boldsymbol\theta$} is the set of parameters and \textbf{\emph{$\boldsymbol b$}} is the set of partitions.
@@
 
-\begin{MPEquation}[!ht]
 \begin{equation}
 P(\boldsymbol b | \boldsymbol A) = \frac{P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)P(\boldsymbol\theta, \boldsymbol b)}{P(\boldsymbol A)}
+\label{Equation_posterior}
 \end{equation}
-\label{Equation_posterior}
-\end{MPEquation}
-and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length
 
-\begin{MPEquation}[!ht]
+and inference is performed by maximising the posterior probability. The numerator in Eq. \ref{Equation_posterior} can be rewritten exponentiating the description length
+
 \begin{equation}
 \Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
+\label{Equation_entropy}
 \end{equation}
-\label{Equation_entropy}
-\end{MPEquation}
 
@@
 This approach identifies cell groups in a statistical robust way and, moreover, it is able to determine the likelihood of the grouping, thus allowing model selection.
-In addition, it is possible to measure the confidence of assignment to groups.
-\textcolor{red}{We show that such information may be exploited to perfect the notion of cell groups and the identification of markers.}
+In addition, it is possible to measure the confidence of assignment to groups, a measure that can be exploited in various analysis tasks.
 
@@
 
-\texttt{schist} is a convenient wrapper to the \emph{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}.
+\texttt{schist} is a convenient wrapper to the \texttt{graph-tool} python library, written in python and designed to be used with \texttt{scanpy}.
 The most prominent function is \texttt{schist.inference.nested\_model()} which takes a \texttt{AnnData} object as input and fits a nested Stochastic Block Model on the \emph{k}NN graph built with \texttt{scanpy} functions (\emph{e.g.} \texttt{scanpy.tools.neighbors()}).
@@
 \texttt{schist} then annotates cells in the data object with all the groups found at each level of a hierarchy.
-Given the large size of the NN graph, it is possible that a single solution represents local minima of the fitting process.
+Given the large size of the NN graph in real-world experiments, it is possible that a single solution represents local minima of the fitting process.
 In addition, it is possible that multiple solutions are equally acceptable to represent the graph partitioning and a better description is given by the consensus over such solutions \cite{Peixoto_2021_consensus}.
@@
 In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
+
 Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
@@
 
-\subsection*{\texttt{schist} does not return artificial cell groupings}
+\subsection*{Absence of information is correctly handled by nSBM}
 
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.png}
 \caption[]{nSBM results for randomised data.
@@
 
-\textcolor{red}{TO BE UPDATED, possibly with some simulation study}
-
-To benchmark \texttt{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data.
-In particular, we used a mixture of 5 cell lines profiled with Chromium 10x platform.
+To benchmark \texttt{schist}, we tested it on scRNA-seq mixology data \cite{Tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data.
+In particular, we used the mixture of 5 cell lines profiled with Chromium 10x platform.
 At a first evaluation of the UMAP embedding, all lines appear well separated.
@@
 \ref{FigureTian}A).
-Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.854$), with a further split in H2228 cell line (Fig.
-\ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.565$).
-
- Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph.
-We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level.
-We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours.
-In general, higher number of components and neighbours has a positive impact on the performance (Fig.
-\ref{Figure1}B).
-Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours.
-If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (Fig.
-S1).
-Running MCMC algorithm recovers the performances of the majority of the configurations (Fig.
-S2).
+Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.829$), with a further split in H2228 cell line (Fig.
+\ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.549$, Fig.
+\ref{FigureTian}C).
+The nSBM solution correctly identifies H1975 groups as a single entity at level 1 of the hierarchy.
+We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes.
+Given the ground truth, the cell lines, SCCAF is able to assess the maximal accuracy that can be achieved in the dataset (0.992).
+When trained with this target accuracy, SCCAF precisely reconstructs the original cell line annotations starting from \texttt{schist} partitions with high accuracy (Fig.
+\ref{FigureTian}D).
+When Leiden partitions are set as input, SCCAF merges H2228 and HCC827 cells into a single cluster and keeps H1975 cells split into two groups (Fig.
+\ref{FigureTian}E), highlighting potential limitations of this approach.
 
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,width=0.5\textheight]{FigureTian.png}
-\caption[]{\emph{schist} applied to scRNA-seq mixology data.
-(A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper.
-A small number of H1975 cells are found in HCC827 and H838 clusters.
-(B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs.
-We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}.
-Adjusted Rand Index between the actual cell lines and the identified groups is shown.
-Darker blue indicates higher concordance between the model and the ground truth.
-(C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below).
-The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{FigureTian}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.png}
+\caption[]{Analysis of a mixture of 5 known cell lines.
+(A) UMAP embedding colored and labeled by cell line identity.
+H1975 cells (orange) show two distinct groups reflecting an internal heterogeneity.
+(B) }\label{FigureTian}
 \end{figure}
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.7\textheight]{Figure_Hemato_Lineage.png}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Hemato_Lineage.png}
 \caption[]{Analysis of hematopoietic differentiation.
@@
 
-\textcolor{red}{TO BE DONE.}.
+\textcolor{red}{TO BE WRITTEN} Used Splatter [ref], more sensitivity, less precision (Fig. \ref{Figure_DE})
 
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_Label_Transfer.png}
+\includegraphics[keepaspectratio,width=0.9\textwidth,height=\textheight]{Figure_Label_Transfer.png}
 \caption[]{Label transfer using SBM.

### 2021-04-20 manuscript.tex
Marker genes done

@@
 \ref{FigureTian}E), highlighting potential limitations of this approach.
+In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (Fig.
+\ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig.
+\ref{Tian_NSBM0}B).
 
@@
 H1975 cells (orange) show two distinct groups reflecting an internal heterogeneity.
-(B) }\label{FigureTian}
+(B) UMAP embedding showing cells colored by level 1 of the hierarchy proposed by the nested Stochastic Block Model.
+(C) UMAP embedding showing cells colored according to the Leiden method at resolution $\gamma = 1$.
+(D) UMAP embedding colored by the classification made by SCCAF when partitions in (B) are used.
+(E) UMAP embedding colored by the classification made by SCCAF when partitions in (C) are used.}\label{FigureTian}
 \end{figure}
 
-%Analysis of the nSBM hierarchy reveals that five levels are needed to describe the experiment (Fig.
-\ref{Figure1}C, upper panel), with level 2 properly catching the cell identity ($ARI = 0.977$).
-In addition to the five major groups, observe two small groups, summing to 11 cells, that were merged to HCC827 and H838 at hierarchy level 3.
-Interestingly, these groups are enriched in cells whose identity was reassigned from H1975 to H838 or HCC827 in the original paper using Demuxlet \cite{kang_2018}, indicating that nSBM was able to recognise peculiar properties and isolate them.
-It may be worth mention that the second best ranked group by cell affinity for these 11 cells is the correct group assigned in the original paper, except for a single cell assigned to H2228.
-%As a high separation between cell lines is observable, optimisation of modularity by Leiden algorithm is also able to identify cell identities with high precision, given that a proper resolution threshold is set (Fig.
-\ref{Figure1}C, lower panel); we found that when resolution is set to 0.05 the cell lines are properly separated ($ARI = 0.975$), with the exception of the above mentioned cells.
-
-%These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter.
-These data also hint at the possibility to identify rare cell types in larger populations.
-
-\subsection*{Comparison with other methods}
-
-\textcolor{red}{TO BE FILLED with SSCAF?}
+In all, these data support the suitability of \texttt{schist}, hence of Nested Stochastic Block Models, for cell group identification in single cell studies.
 
@@
 Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
-To show this effect we took advantage of public spatial RNA dataset of murine cerebellum profiled with 10X Visium HD technology [\textcolor{red}{REF}], as provided by the recently introduced package \texttt{SquidPy} \cite{Palla_Theis_2021}, for which we chose to stick to the given tissue annotation by the package authors.
-At default resolution, modularity optimization resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig.
+The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.
+
+To show this we took advantage of public spatial RNA dataset of a coronal section of murine brain tissue profiled with 10X Visium H\&E technology \cite{Gracia_Villacampa_2020}, as provided by the recently introduced package \texttt{SquidPy} \cite{Palla_Theis_2021}.
+We chose to stick to the given tissue annotation by the package authors.
+At default resolution, Leiden clustering resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig.
 \ref{Figure_Visium}).
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Visium_Spatial.png}
-\caption[]{Analysis of spatial transcriptomics of cerebellum by Visium HD.
+\caption[]{Analysis of spatial transcriptomics of a coronal section of mouse brain by Visium H\&E.
 In the first panel, original tissue annotation is given.
@@
 
-In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameters.
+In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameter.
 We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019} (Fig.
@@
 Cells are colored according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A).
-In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure_Hemato_Lineage}
+In subsequent panels, cells are colored using a signature of Erythroid lineage (B), Monocytes (C) or Neutrophils (D).}\label{Figure_Hemato_Lineage}
 \end{figure}
@@
 Not only the hierarchy is not conserved, but also very different cell types may be mixed in spurious clusters.
-By using SBM, \texttt{schist} is able to represent hierarchical relations in appropriate way.
+By using nSBM, \texttt{schist} is able to represent hierarchical relations in appropriate way.
 Moreover, the hierarchy appears to be more robust in aggregating different cell types at coarser scales.
 
-%\subsection*{Hierarchic model constrains differential gene tests}
-%Current approaches to test for gene markers in scRNA-seq data rely on pairwise or one-vs-rest comparisons among cell groups.
-This strategy assumes that cell groups are independent which, in general, may not be true.
-A notable exception is the recently introduced treeclimbR \cite{huang_2020} which has been specifically developed to perform differential tests over a tree structure.
-We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.
+\subsection*{Cell marginals have little impact on detection of marker genes}
 
-\subsection*{Cell marginals improve detection of marker genes}
-
-\textcolor{red}{TO BE WRITTEN} Used Splatter [ref], more sensitivity, less precision (Fig. \ref{Figure_DE})
+The default \texttt{schist} results include a matrix (\emph{cell marginals}) that is defined as the marginal posterior probabilities of each cell to be assigned to each group.
+These are calculated according to the consensus partition estimated by \texttt{graph-tool} using an instance of the \texttt{PartitionModeState} class.
+These marginals may be helpful when the experimenter wants to isolate cells that have been assigned with strong evidence to cell groups, or for general quality control assessment.
+We instead sought to investigate if they can be used as continuous covariates when testing for differentially expressed genes among cell groups.
+To this end, we took advantage of the \texttt{Splatter} package \cite{Zappia_Oshlack_2017} to simulate 6 cell populations at different levels of noise (biological coefficient of variation, $BCV$, ranging from 0.2 to 2) and probabilities to have differentially expressed genes within a group ($p(DE)$, ranging from 0.1 to 1).
+For each cell group we built two Generalized Linear Models, one with dummy variables recapitulating the standard \emph{t}-test that is performed by \texttt{scanpy} (\emph{i.e.} $\sim1 + group$) and one using the marginal (\emph{i.e.} $\sim1 + group\_marginal$).
+In general we did not observe substantial changes between the two models (Fig.
+\ref{Figure_DE}).
+Using marginals increases the amount of positive tests (at $FDR<0.01$) and the sensitivity of the method, especially when the data are noisy, but at the cost of the reduction of recall due to higher rate of false positives.
+In all, the utility of cell marginals to this task remains dubious.
 

### 2021-04-20 supplementary.tex
use marginals for gene xp
Marker genes done

@@
 
-%\begin{figure}[htbp]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S1.png}
-%\caption[]{Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors.
-Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis).
-Both the parameters influence the sparseness of the graph.}\label{MPFigure:F56BBDC3-0F0A-4558-8BC2-839E91250DB4}
-%\end{figure}
-
+\begin{figure}[htbp]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Tian_NSBM0.png}
+\caption[]{Analysis of a mixture of 5 known cell lines.
+(A) UMAP  embedding of cell lines from the sc-mixology experiment colored by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding colored by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
+\end{figure}
+\clearpage
 %\begin{figure}[htbp]
@@
 \end{figure}
+\clearpage
 
@@
 \end{figure}
+\clearpage
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_DE.png}
+\caption[]{\textcolor{red} WRITE ONE. }\label{Figure_DE}
+\end{figure}
+\clearpage
 

### 2021-04-20 manuscript.tex
start something on optimal level

@@
 \ref{FigureTian}E), highlighting potential limitations of this approach.
-In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (Fig.
+In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig.
 \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig.
@@
 
-The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme.
+The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself.
 Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data.
 We sought to exploit this property to address the task of annotating cells according to a reference sample.
-To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies.
-We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 and MARS-seq.
+To this end we analyzed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies.
+We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 or MARS-seq.
 These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.
+
 After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph.
-In each merged dataset, we retained cell type annotations for 10XV3 cells, while we assigned a 'Unknown' label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2).
-We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'.
+In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2).
+We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown".
 Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood.
-The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their unannotated label if and only if there is not enough evidence to associate them to another group.
-We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in Scanpy also based on \emph{k}NN classification.
+The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their "Unknown" label if and only if there is not enough evidence to associate them to another group.
+We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings.
 Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig.
@@
 When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabeled cells.
-This effect is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243).
+This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243).
 Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641).
 Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely).
-Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity was indeed the appropriate one (Fig.
+Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity ranking was indeed the appropriate one (Fig.
 \ref{hist_label_second}).
@@
 
-\subsection*{Choice of the optimal level}
+\subsection*{Choice of an optimal hierarchy level}
 
-\textcolor{red}{TO BE DONE. Discuss modularity, but also eigenanalysis of random matrix}.
+The Nested Stochastic Block Model, hence \texttt{schist}, fits a hierarchical model of communities into a graph.
+When it comes to analysis of single cell data, it means that the cells are best described by the hierarchy itself and that cells \emph{can} be grouped consistently at each level of the tree.
+In addition, the size of groups at the deepest level scales as $O(N/\log{N})$ \cite{peixoto_2014_h}, where $N$ is the number of cells.
+Given the current throughput in single cell experiments ($\sim$10k cells), the number of groups becomes intractable.
+For this reason, in most of single cell experiments, it is preferable to identify an optimal level of the hierarchy that best resembles the cell properties at the scale they can be validated.
 
-\subsection*{PPBM?}
+A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution.
+According to this strategy, the optimal level $i$ is the one that minimizes the number of partitions and $\hat{k}$:
 
-\textcolor{red}{TO BE DONE. Maybe this is the right place to introduce the choice of the appropriate hierarchy level}
+\begin{equation}
+i = \underset{x}{\mathrm{argmin}} |B_x - \hat{k}|
+\label{Equation_RMT}
+\end{equation}
+
+where $B_x$ is the number of non empty partitions at level $x$.
+
+An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels.
+While nSBM does not optimize the graph modularity, we observed that this tends to be maximal for the level better describing known cell populations.
+
+%\subsection*{PPBM?}
+
+%\textcolor{red}{TO BE DONE. Maybe this is the right place to introduce the choice of the appropriate hierarchy level}
 
@@
  \hline\hline
-% sc-mixology \cite{tian_2019} & 860 & 01:11 & 00:03 & 0.884 \\ 
- %\hline
+ sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:08 & 00:55 & 01:23 \\ 
+ \hline
  Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 

### 2021-04-20 supplementary.tex
start something on optimal level

@@
 \includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_DE.png}
-\caption[]{\textcolor{red} WRITE ONE. }\label{Figure_DE}
+\caption[]{Results of differential gene expression with or without marginals.
+The heatmaps show the difference in sensitivity (left) or recall (right) of statistical testing for differential genes among cell groups when marginals are included or not in the model.
+Tests have been conducted on simulated data with different levels of noise (BCV) or probability of differential expression (p(DE))}\label{Figure_DE}
 \end{figure}

### 2021-04-20 manuscript.tex
again optimal level

@@
 A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution.
-According to this strategy, the optimal level $i$ is the one that minimizes the number of partitions and $\hat{k}$:
+According to this strategy, the optimal level $i^k$ is the one that minimizes the number of partitions and $\hat{k}$:
 
 \begin{equation}
-i = \underset{x}{\mathrm{argmin}} |B_x - \hat{k}|
+i^k = \underset{x}{\mathrm{argmin}} |B_x - \hat{k}|
 \label{Equation_RMT}
@@
 An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels.
-While nSBM does not optimize the graph modularity, we observed that this tends to be maximal for the level better describing known cell populations.
+While nSBM does not optimize the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is
+
+\begin{equation}
+i^Q = \underset{x}{\mathrm{argmax}} |Q_x|
+\label{Equation_MODmax}
+\end{equation}
+
+Where $Q_x$ is modularity at level $x$.
+We tabulated values arising from both the approaches in Table \ref{table_optimal}, together with the $i^D$ that is defined similarly to Eq.
+\ref{Equation_RMT} using the number of known annotated groups $D$ in place of $\hat{k}$
+
+\begin{table}[h!]
+\centering
+ \begin{tabular}{|| l c c c c c c c c c ||}
+ \hline
+ \textbf{Dataset} & \textbf{Cells} & \textbf{D} & \textbf{D} & \textbf{nD} & \textbf{$\hat{k}$} & \textbf{$i$} & \textbf{$B_i$} & \textbf{Annotated groups} & \textbf{Annotated groups} \\ [0.5ex]
+
+ \hline\hline
+% sc-mixology \cite{Tian_2019} & 860 & 5 & 00:08 & 00:55 & 01:23 \\
+
+\end{tabular}
+\caption{foo bar}
+\label{table_optimal}
+\end{table}
 
@@
 \end{tabular}
-\caption{Time required to minimise the nSBM using the default minimization method compared to the greedy merge-split MCMC.
-Times are expressed in mm:ss.
-Partition overlap measures concordance between the two models over the full hierarchy.
-Timing is the average after 3 initialisations.}
+\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. Times are expressed in mm:ss. }
 \label{Table1}

### 2021-04-20 manuscript.tex
something on timing

@@
 
-Where $Q_x$ is modularity at level $x$.
-We tabulated values arising from both the approaches in Table \ref{table_optimal}, together with the $i^D$ that is defined similarly to Eq.
-\ref{Equation_RMT} using the number of known annotated groups $D$ in place of $\hat{k}$
+Where $Q_x$ is modularity at level $x$. We tabulated values arising from both the approaches in Table \ref{table_optimal}
 
@@
 \centering
- \begin{tabular}{|| l c c c c c c c c c ||}
+ \begin{tabular}{|| l c c c c c c c ||}
  \hline
- \textbf{Dataset} & \textbf{Cells} & \textbf{D} & \textbf{D} & \textbf{nD} & \textbf{$\hat{k}$} & \textbf{$i$} & \textbf{$B_i$} & \textbf{Annotated groups} & \textbf{Annotated groups} \\ [0.5ex]
+ \textbf{Dataset} & \textbf{Cells} & \textbf{$D$}  & \textbf{$\hat{k}$} & \textbf{$i^k$} & \textbf{$B_k$} & \textbf{$i^Q$} & \textbf{$B_Q$} \\ [0.5ex]
 
  \hline\hline
-% sc-mixology \cite{Tian_2019} & 860 & 5 & 00:08 & 00:55 & 01:23 \\
-
+ sc-mixology \cite{Tian_2019} & 860 & 5  & 21 & 1 & 6 & 1 & 6 \\ 
+\hline
+ Chromium 10x \cite{mereu_2020} & 1523 & 8  & 43 & 0 & 58 & 1 & 13 \\
+ \hline
+ Quartz-seq2 \cite{mereu_2020} & 1266 & 8  & 37 & 0 & 62 & 1 & 12 \\
+ \hline
+ MARS-seq \cite{mereu_2020} & 1401 & 9  & 9 & 1 & 16 & 1 & 16 \\
+ \hline
+ iCELL8 \cite{mereu_2020} & 1830 & 9  & 20 & 1 & 21 & 2 & 6 \\
+ \hline
+ Mouse brain \cite{Gracia_Villacampa_2020} & 2688 & 15  & 8 & 2 & 8 & 1 & 23 \\
+ \hline
+ Planaria \cite{plass_2018} & 21612 & $51^\ast$ & 8 & 2 & 8 & 1 & 23 \\
+ \hline
 \end{tabular}
-\caption{foo bar}
+\caption{Selection of the optimal level in the nSBM hierarchy.
+For each dataset we report the number of groups $D$ that were given by the authors.
+The optimal level selection should recover a number of groups in the order of magnitude of $D$.
+Value of $D$ in Planaria dataset is derived from manual curation of Louvain clustering.
+$\hat{k}$:  number of groups according to RMT, $i^k$: level selected according to RMT criterion, $B_k$: number of partitions at level $i^k$, $i^Q$: level at which modularity is maximal, $B_Q$: number of groups at level $i^Q$}
 \label{table_optimal}
@@
 
+\textcolor{red}{Add conclusions for this paragraph}
 %\subsection*{PPBM?}
@@
 
-Minimisation of the nSBM is a process that may require a large amount of computational resources.
-The analysis of a relatively small scRNA-seq dataset, such as the ones in \cite{mereu_2020}, may require several minutes to be processed.
-This could be a serious limitation to the adoption of nSBM in the analysis of single cell data, especially because several parameters should be tested.
-To overcome this limitation, it was suggested to let a greedy merge-split MCMC algoritthm \cite{peixoto_2020} to explore the solutions and stop iterations when the difference in entropy is below a defined threshold.
-We tested this approach on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM) and compared to the default approach.
-Results are reported in Table \ref{Table1}.
-The merge-split algorithm greatly reduces the time needed to propose the final model.
-In addition, the partitions found are largely overlapping the ones found by the default approach.
+Minimisation of the nSBM is a process that requires a large amount of computational resources.
+While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges.
+Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged.
+To give a reference, we collected runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM).
+Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more and $\sim180\times$ times more, at worst (possibly due to limited memory for large networks).
 
@@
  \hline
- Paul15 \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
+ Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
  \hline
@@
 \caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. Times are expressed in mm:ss. }
-\label{Table1}
+\label{Table_runtime}
 \end{table}

### 2021-04-20 manuscript.tex
something on timing

@@
 To give a reference, we collected runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM).
-Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more and $\sim180\times$ times more, at worst (possibly due to limited memory for large networks).
+Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more, and $\sim180\times$  at worst (possibly due to limited memory for large networks).
+
+\textcolor{red}{Add conclusions for this paragraph}
 

### 2021-04-21 manuscript.tex
RMT for Planaria
only conclusions left

@@
 
-Where $Q_x$ is modularity at level $x$. We tabulated values arising from both the approaches in Table \ref{table_optimal}
+Where $Q_x$ is modularity at level $x$. We collected values arising from both the approaches for some datasets used in this work (Table \ref{table_optimal})
 
@@
  \hline
- Planaria \cite{plass_2018} & 21612 & $51^\ast$ & 8 & 2 & 8 & 1 & 23 \\
+ Planaria \cite{plass_2018} & 21612 & $51^\ast$ & 34 & 2 & 22 & 3 & 10 \\
  \hline
@@
 
-\textcolor{red}{Add conclusions for this paragraph}
+As expected, the larger the network, the higher the optimal level.
+For relatively small datasets (\emph{i.e.} less than 10k cells), the first level of the hierarchy contains a number of groups in line with how many observable populations are.
 %\subsection*{PPBM?}
@@
 Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged.
-To give a reference, we collected runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM).
+To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM).
 Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more, and $\sim180\times$  at worst (possibly due to limited memory for large networks).
-
-\textcolor{red}{Add conclusions for this paragraph}
+A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes.
+The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.
 
@@
  \hline
- \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{NSBM} \\ [0.5ex] 
+ \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
  \hline\hline
@@
  \hline
-% 4 & 545 & 18744 & 7560 \\
-% \hline
-% 5 & 88 & 788 & 6344 \\ [1ex] 
-% \hline
 \end{tabular}
@@
 \section*{Conclusions}
-\textcolor{red}{TO BE UPDATED}
 

### 2021-04-21 manuscript.tex
new section
boh

@@
 
+\subsection*{Cell marginals can be used to assess the data quality}
+\textcolor{red}{TO BE DONE, is this true anyway?}
+
 \subsection*{Cell marginals have little impact on detection of marker genes}
@@
 In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.
-Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity.
-Lastly, a statistical way to evaluate models is made available, allowing for reliable model selection.
-This last capability has the obvious advantage that the choice of parameters, hence the definition of cell clusters, could be conditioned to an evaluation metric which is robust and easy to understand (\emph{i.e.} the model entropy).
+\textcolor{red}{Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity}.
+
+The Bayesian formulation of Stochastic Block Models provides the possibility to perform inference on a graph for any partition configuration, thus allowing reliable model selection using an interpretable measure, entropy.
+We exploited this property to perform label transfer with high accuracy and with the possibility to discard cells with unreliable assignments.
+In all, \texttt{schist} facilitates the adoption of nSBM by the bioinformatics community and exposes a robust framework to perform tasks that go beyond the principled identification of cell clusters.
 
 The major drawback of adopting this strategy is the substantial increase of runtimes.
-According to the developers of \emph{graph-tool}, runtimes are proportional to the number of edges in the neighbourhood graph and while it supports CPU-level parallelisation, a model minimisation is hundreds times slower than the extremely fast Leiden approach.
-Nevertheless, we show that a greedy merge-split MCMC algorithm can overcome this limitation, achieving performances that allow the usage of \emph{schist} on standard desktop hardware to analyse various single cell datasets.
+As observed, model minimisation may be hundred times slower than the extremely fast Leiden approach.
+It should be noted that \texttt{schist} initializes multiple models that are treated by multiple concurrent processes.
+\texttt{graph-tool} itself supports CPU-level parallelization for some of its tasks.
+These optimizations are well suited for clustered computing infrastructure.
+Further development, possibly including GPU-level parallelization, is surely required to accomodate the large size of datasets that are being produced.
 
@@
 
-\textcolor{red}{TO BE UPDATED}
+Unless differently stated, all the analysis were produced using \texttt{scanpy} v1.7.1 \cite{wolf_2018} and \texttt{schist} v0.7.6 and the corresponding dependencies.
+All models were initialized 100 times, here including Leiden partitioning for which we also calculated the consensus partition.
 
 \subsection*{Analysis of Random data}
-Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function.
+
+Data were retrieved in \texttt{scanpy} environment using \texttt{sc.datasets.pbmc3k\_processed()} function.
 The random \emph{k}NN graph was obtained shuffling the node labels of each edge.
-nSBM was performed with 3 initialisations using merge-split MCMC approach.
 UMAP embedding was recomputed after randomisation using the shuffled graph.
@@
 Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}).
-Data were analysed using scanpy v1.4.6 \cite{wolf_2018}.
 Cells with less than 200 genes were excluded, as genes detected in less than 3 cells.
@@
 Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out.
-nSBM was initialised three times
+\emph{k}NN graph was built with default parameters (50 components and 15 nearest neighbors).
+Data were assessed by SCCAF using cell line annotation.
+Mean cross-validated accuracy was set as target for all the models.
+
+\subsection*{Analysis of Visium H\&E data}
+\textcolor{red}{TOBEDONE}
 
@@
 
-Data were retrieved using scanpy's built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbours and diffmap as embedding.
+Data were retrieved using \texttt{scanpy}'s built-in functions and were processed as in \cite{wolf_2019}, except for \emph{k}NN graph built using 30 principal components, 30 neighbours and diffmap as embedding.
 Gene signatures were calculated using the following gene lists
+
 \begin{itemize}
@@
 
-nSBM was completed with 3 initialisations
+\subsection*{Label transfer}
+\textcolor{red}{TO BE DONE}
 

### 2021-04-22 manuscript.tex
Added paragraph on QC

@@
 \subsection*{Cell marginals can be used to assess the data quality}
-\textcolor{red}{TO BE DONE, is this true anyway?}
+
+By computing the consensus among multiple models, \texttt{schist} returns the marginal probability for each cell to belong to a specific cluster at each level of the hierarchy.
+Ideally, all cells should always be assigned with $p=1$ to a cluster.
+When when the uncertainty is maximal, cells are assigned to clusters randomly with $p=1/B_i$, where $B_i$ is the number of groups for the $i$-th level in the hierarchy.
+We sought to check if these probabilities could be interpreted in terms of data quality.
+
+We devised a simple metric, \emph{cell stability}, that is defined by the fraction of levels for which the marginal probability is higher than $1-1/B_i$.
+To do so, we only consider levels with at least two groups, hence excluding the root of the tree.
+We tested this metric on four datasets from \cite{mereu_2020} with different quality levels (iCELL8, MARS-seq, 10XV3 and Quartszeq-2) (Fig.
+\ref{Figure_Stability}).
+By taking a summary metric, \emph{e.g.} the mean $\overline{S}$ or the fraction of cells with $S>0.5$, we observed that it correlates with the data quality (Table \ref{table_stability})
+
+\begin{table}[h!]
+\centering
+ \begin{tabular}{|| l c c ||}
+ \hline
+ \textbf{Dataset} & \textbf{$\overline{S}$} & \textbf{$S > .5$} \\ [0.5ex]
+
+ \hline\hline
+ iCELL8 \cite{mereu_2020} & 0.368 & 0.312  \\
+ \hline
+ MARS-seq \cite{mereu_2020} & 0.579 & 0.536  \\
+ \hline
+ Chromium 10x \cite{mereu_2020} & 0.716 & 0.728  \\
+ \hline
+ Quartz-seq2 \cite{mereu_2020} & 0.705 & 0.739 \\
+ \hline
+\end{tabular}
+\caption{Cell stability as indicator of data quality.
+Table shows summary metrics derived from the Cell Stability calculated for various datasets.
+$\overline{S}$ is the average Cell Stability over all cells, $S>.5$ indicates the fraction of cells with Cell Stability higher than 0.5.}
+\label{table_stability}
+\end{table}
+
+These data suggest that measures of uncertainty of cell clustering can be useful for general quality control assessment.
+In addition to this, we foresee they could be used to isolate cells with specific patterns.
 
@@
 
-The default \texttt{schist} results include a matrix (\emph{cell marginals}) that is defined as the marginal posterior probabilities of each cell to be assigned to each group.
-These are calculated according to the consensus partition estimated by \texttt{graph-tool} using an instance of the \texttt{PartitionModeState} class.
-These marginals may be helpful when the experimenter wants to isolate cells that have been assigned with strong evidence to cell groups, or for general quality control assessment.
-We instead sought to investigate if they can be used as continuous covariates when testing for differentially expressed genes among cell groups.
+We sought to investigate if cell marginals can be used as continuous covariates when testing for differentially expressed genes among cell groups.
 To this end, we took advantage of the \texttt{Splatter} package \cite{Zappia_Oshlack_2017} to simulate 6 cell populations at different levels of noise (biological coefficient of variation, $BCV$, ranging from 0.2 to 2) and probabilities to have differentially expressed genes within a group ($p(DE)$, ranging from 0.1 to 1).
 For each cell group we built two Generalized Linear Models, one with dummy variables recapitulating the standard \emph{t}-test that is performed by \texttt{scanpy} (\emph{i.e.} $\sim1 + group$) and one using the marginal (\emph{i.e.} $\sim1 + group\_marginal$).
-In general we did not observe substantial changes between the two models (Fig.
+In general, we did not observe substantial changes between the two models (Fig.
 \ref{Figure_DE}).
-Using marginals increases the amount of positive tests (at $FDR<0.01$) and the sensitivity of the method, especially when the data are noisy, but at the cost of the reduction of recall due to higher rate of false positives.
+Using marginals slightly increases the amount of positive tests (at $FDR<0.01$) and the sensitivity of the method, especially when the data are noisy, but at the cost of the reduction of recall due to an increased rate of false positives.
 In all, the utility of cell marginals to this task remains dubious.
@@
 To this end we analyzed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies.
-We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 or MARS-seq.
+We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq.
 These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task.
@@
 After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph.
-In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2).
+In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2).
 We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown".
@@
 We compared the accuracy of the  outcome to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in \texttt{scanpy}  based on \emph{k}NN classification of UMAP embeddings.
-Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig.
+Analysis of a well defined dataset, such as Quartz-seq2, reveals that the three approaches are equally good in classifying unknown cells (Fig.
 \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927.
@@
 \caption[]{Label transfer using SBM.
-Annotated data from Chromium 10X (A) and  MARS-seq (B) were merged using Harmony.
-MARS-seq cells were relabeled according to the \emph{k}NN graph (C) or by exploiting the Stochastic Block Model (D).
-}\label{Figure_Label_Transfer}
+The first line reports UMAP embeddings for datasets profiled with Chromium 10X V3, Quartz-seq2 and  MARS-seq, each annotated by known cell types.
+Quartz-seq2 and MARS-seq were reannotated using \emph{k}NN method, \texttt{sc.tl.ingest() } or \texttt{schist}.
+The accuracy of each label transfer task is reported above the corresponding UMAP.} \label{Figure_Label_Transfer}
 \end{figure}

### 2021-04-23 manuscript.tex
mod conclusions
update runtimes by Leonardo

@@
 Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged.
-To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM).
-Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more, and $\sim180\times$  at worst (possibly due to limited memory for large networks).
+To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (\textcolor{red}{dual core 1.6 GHz i5 processor, 16 GB RAM}).
+Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst.
 A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes.
@@
  \hline\hline
- sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:08 & 00:55 & 01:23 \\ 
+ sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
  \hline
- Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 
+ Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
  \hline
- Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:16 & 00:30 & 02:10 \\
+ MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
  \hline
- MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:24 & 01:10 & 06:11 \\
+ iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
  \hline
- iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:34 & 01:31 & 09:06 \\
+ Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
  \hline
- Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
+ Islet \textcolor{red}{??} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
  \hline
- Planaria \cite{plass_2018} & 21612 & 173667 & 06:32 & 34:12 & 3:17:08 \\
+ Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
+ \hline
+ Zeisel Cortex \textcolor{red}{??} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
+ \hline
+ \textcolor{red}{Pancreas velocity} \textcolor{red}{??} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
+ \hline
+ Baron Pancreas \textcolor{red}{??} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
+ \hline
+ Airzani Liver \textcolor{red}{??} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
+ \hline
+ Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
  \hline
 \end{tabular}
-\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. Times are expressed in mm:ss. }
+\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets.
+All approaches fit 100 models.
+Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}.
+Times are expressed in mm:ss.
+}
 \label{Table_runtime}
@@
 In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.
-\textcolor{red}{Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity}.
 

### 2021-04-26 manuscript.tex
Update references

@@
+% !TEX encoding = UTF-8 Unicode
 \documentclass[10pt]{article}
@@
 \makeatother
+
+\inputencoding{utf8}
 
@@
 Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged.
-To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (\textcolor{red}{dual core 1.6 GHz i5 processor, 16 GB RAM}).
+To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM).
 Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst.
@@
  Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
- \hline
- Islet \textcolor{red}{??} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
+% \hline
+% Islets \cite{Segerstolpe_2016} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
  \hline
@@
  \hline
- Zeisel Cortex \textcolor{red}{??} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
+ Mouse Cortex \cite{Zeisel_2015} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
  \hline
- \textcolor{red}{Pancreas velocity} \textcolor{red}{??} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
+ Endocrinogenesis \cite{Bastidas_Ponce_2019} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
  \hline
- Baron Pancreas \textcolor{red}{??} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
+ Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
  \hline
- Airzani Liver \textcolor{red}{??} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
+ Airzani Liver \cite{Aizarani_2019} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
  \hline
@@
 Unless differently stated, all the analysis were produced using \texttt{scanpy} v1.7.1 \cite{wolf_2018} and \texttt{schist} v0.7.6 and the corresponding dependencies.
-All models were initialized 100 times, here including Leiden partitioning for which we also calculated the consensus partition.
+All models were initialized 100 times, herein including Leiden partitioning for which we also calculated the consensus partition.
 

### 2021-04-26 manuscript.tex
more methods

@@
 % !TEX encoding = UTF-8 Unicode
+
 \documentclass[10pt]{article}
@@
 \subsection*{Analysis of Visium H\&E data}
-\textcolor{red}{TOBEDONE}
+Data were retrieved using \texttt{squidpy.datasets.visium\_hne\_adata()} built-in function, without further processing.
+Leiden clustering was performed using \texttt{schist.inference.leiden()} function, allowing for 100 initializations, with resolutions $\gamma = 1$ and $\gamma = 0.5$.
 
@@
 
-\subsection*{Label transfer}
-\textcolor{red}{TO BE DONE}
-
-\subsection*{Analysis of cluster consistency}
-
+\subsection*{Processing of PBMC data from various platforms}
 Count matrices were downloaded from GEO using the following accession numbers: GSE133535 (Chromium 10Xv3), GSE133543 (Quartz-seq2), GSE133542 (MARS-seq) and GSE133541 (iCELL8).
@@
 Neighbourhood graph was built using 30 principal components and 20 neighbours.
-nSBM was completed with 3 initialisations.
+
+\subsection*{Label transfer}
+
+Processed data for MARS-seq or Quart-seq2 platforms were merged to data for 10X V3.
+Merged data were then processed using Harmony \cite{Korsunsky_2019} by the \texttt{scanpy.external.pp.harmony\_integrate()} function with default parameters.
+Cells not belonging to the 10X data were assigned an "Unknown" label.
+We calculated cell affinity to each annotation label using \texttt{schist.tl.calculated\_affinity()} function.
+We assigned the most affine annotation only to "Unknown" cells.
+For \emph{k}NN-based procedure, we built a \emph{k}NN graph on the merged data using \texttt{pynndescent } library on the 10XV3 subset of cells in the merged data, then we assigned "Unknown" cells to the closest entry in the graph.
+Assignment by \texttt{scanpy.tl.ingest()} was performed using default parameters.
 

### 2021-04-26 manuscript.tex
removed the DE part

@@
 
-\subsection*{Cell marginals have little impact on detection of marker genes}
+%\subsection*{Cell marginals have little impact on detection of marker genes}
 
-We sought to investigate if cell marginals can be used as continuous covariates when testing for differentially expressed genes among cell groups.
+%We sought to investigate if cell marginals can be used as continuous covariates when testing for differentially expressed genes among cell groups.
 To this end, we took advantage of the \texttt{Splatter} package \cite{Zappia_Oshlack_2017} to simulate 6 cell populations at different levels of noise (biological coefficient of variation, $BCV$, ranging from 0.2 to 2) and probabilities to have differentially expressed genes within a group ($p(DE)$, ranging from 0.1 to 1).

### 2021-04-26 supplementary.tex
Added paragraph on QC
removed the DE part

--- 2021-04-20 supplementary.tex
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_DE.png}
-\caption[]{\textcolor{red} WRITE ONE. }\label{Figure_DE}
+\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_Stability.png}
+\caption[]{Cell Stability. UMAP embeddings of PBMC data profiled with four different technologies ranked by their quality. Cells are colored by the Cell Stability metric.}\label{Figure_Stability}
 \end{figure}
 \clearpage
+
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_DE.png}
+%\caption[]{Results of differential gene expression with or without marginals.
+The heatmaps show the difference in sensitivity (left) or recall (right) of statistical testing for differential genes among cell groups when marginals are included or not in the model.
+Tests have been conducted on simulated data with different levels of noise (BCV) or probability of differential expression (p(DE))}\label{Figure_DE}
+%\end{figure}
+%\clearpage
 
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{hist_label_second.png}
-\caption[]{TO BE WRITTEN}\label{hist_label_second}
+\caption[]{Distribution of second choices in label transfer.
+For each dataset in which we assessed accuracy of label transfer by SMB, we collect the rank of affinities of the correct class when an assignment could not be performed (\emph{i.e.} the highest affinity was for the "Unknonwn" label)  }\label{hist_label_second}
 \end{figure}

### 2021-04-27 manuscript.tex
Final rendering?
Update manuscript.tex

@@
 Ideally, all cells should always be assigned with $p=1$ to a cluster.
-When when the uncertainty is maximal, cells are assigned to clusters randomly with $p=1/B_i$, where $B_i$ is the number of groups for the $i$-th level in the hierarchy.
+When the uncertainty is maximal, cells are assigned to clusters randomly with $p=1/B_i$, where $B_i$ is the number of groups for the $i$-th level in the hierarchy.
 We sought to check if these probabilities could be interpreted in terms of data quality.
@@
  \hline
- Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
+ Baron Pancreas \cite{Baron_2016} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
  \hline

### 2021-04-27 manuscript.tex
Update manuscript.tex

@@
 \ref{Figure_Stability}).
-By taking a summary metric, \emph{e.g.} the mean $\overline{S}$ or the fraction of cells with $S>0.5$, we observed that it correlates with the data quality (Table \ref{table_stability})
+By taking a summary metric, \emph{e.g.} the mean $\overline{S}$ or the fraction of cells with $S>0.5$, we observed that it correlates with the data quality (Table \ref{table_stability}).
 

### 2021-04-27 manuscript.tex
Update manuscript.tex

@@
 We also would like to thank all people at COSR, in particular Giovanni Tonon and Paolo Provero.
-This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC
+This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC.
 

### 2021-04-27 supplementary.tex
Update supplementary.tex

@@
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
-Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.} groups 6 and 8 in the Erythroid branch at resolution $\gamma=1$ are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$)}\label{Figure_Paul15_Leiden_r}
+Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.} groups 6 and 8 in the Erythroid branch at resolution $\gamma=1$ are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$).}\label{Figure_Paul15_Leiden_r}
 \end{figure}

### 2021-04-27 manuscript.tex
Louvain limits update
title suggestion 1

@@
 Arguably, the biggest utility of NN structure is the possibility to identify cell groups by partitioning the graph into communities; this is typically achieved using the Louvain method \cite{blondel_2008}, a fast algorithm for optimisation of graph modularity.
-While fast, this method does not guarantee that small communities in large networks are well defined.
+While fast, this method does not guarantee the identification of internally connected communities.
 To overcome its limits, a more recent approach, the Leiden algorithm \cite{traag_2019}, has been implemented and it has been quickly adopted in the analysis of single cell data, for example by \texttt{scanpy} \cite{wolf_2018} and \texttt{PhenoGraph} \cite{levine_2015}.
@@
 
-\subsection*{Hierarchy modelling is accurate in defining biological properties}
+\subsection*{Hierarchy modeling complies with biological properties}
 

### 2021-04-27 manuscript.tex
Modified abstract

@@
 Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system.
-The definition of the  properties of single cells is the primary endpoint of such analysis, nevertheless cells are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation.
+The definition of the  properties of single cells is the primary endpoint of such analysis, cells are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation.
 Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity.
 In this paper we propose an alternative and principled solution to this problem, based on Stochastic Block Models.
-We show that our approach is not only able to correctly identify cell groups, but also adds the possibility to perform inference on cell identities \textcolor{red}{SOMETHING MORE?
-Damn, I always struggle with abstracts}.
+We show that such approach not only is suitable for identification of cell groups, it also provides a solid framework to perform other relevant tasks in single cell analysis, such as label transfer.
+To encourage the use of Stochastic Block Models, we developed a python library, \texttt{schist}, that is compatible with the popular \texttt{scanpy} framework.
 
@@
 
-We developed \texttt{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \texttt{scanpy}, to facilitate the adoption of stochastic block models in single-cell analysis.
+We developed \texttt{schist} (\href{https://github.com/dawe/schist}{https:/\slash github.com\slash dawe\slash schist}), a python library compatible with \texttt{scanpy}, to facilitate the adoption of Stochastic Block Models in single-cell analysis.
 

### 2021-04-27 supplementary.tex
added figure S6

@@
 \usepackage{authblk}
+\usepackage{xr}
+\externaldocument{manuscript}
 
@@
 
-%\begin{figure}[htbp]
-%\centering
-%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S5.png}
-%\caption[]{Adjusted Rand Index between cell clusters and cell type annotation filtering data at different cutoffs of consistency.
-Dot size is proportional to the number of cells remaining after filtering.}\label{MPFigure:5B291ABA-427A-42BE-BC26-38E626B1B088}
-%\end{figure}
+\begin{figure}[htbp]
+\centering
+\includegraphics[keepaspectratio,width=\textwidth,height=1\textheight]{FigureOptimalLevel.png}
+\caption[]{Choice of the optimal hierarchy level.
+Datasets analyzed for choice of the most relevant hierarchy level are represented here, from top to bottom in the same order as in Table \ref{table_optimal}.
+For each dataset we report the UMAP embedding coloured by the annotation given in the corresponding manuscript, the profile of modularity $Q$ at different level of the hierarchy and the UMAP embeddings coloured by the level according to $i^k$ or $i^Q$.
+The dashed line in modularity plots represents the modularity calculated using the annotation from the corresponding manuscripts.}\label{FigureOptimalLevel}
+\end{figure}
 

### 2021-04-27 manuscript.tex
added figure S6
Final retouch

@@
 As the popularity of single cell analysis frameworks \texttt{Seurat} \cite{butler_2018} and \texttt{scanpy} \cite{wolf_2018} raised, methods based instead on graph partitioning became the \emph{de facto} standards.
-Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbors, \emph{s}NN).
+Such methods require the construction of a cell neighbourhood graph (\emph{e.g.} by \emph{k }Nearest Neighbours, \emph{k}NN, or shared Nearest Neighbours, \emph{s}NN).
 Encoding cell-to-cell similarities into graphs has practical advantages beyond clustering, as many algorithms for graph analysis can be applied and interpreted in a biological way.
@@
 
-An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organized into communities \cite{holland_1983}.
+An alternative solution to community detection is the Stochastic Block Model, a generative model for graphs organised into communities \cite{holland_1983}.
 In this scenario, identification of cell groups requires the estimation of the proper parameters underlying the observed NN graph.
@@
 
-so that inference is performed by minimizing the information required to describe the data (Occam's razor); \texttt{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
+so that inference is performed by minimising the information required to describe the data (Occam's razor); \texttt{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}.
 SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}.
@@
 To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}).
-In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximization of modularity.
+In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.
 
@@
 \caption[]{Analysis of a mixture of 5 known cell lines.
-(A) UMAP embedding colored and labeled by cell line identity.
+(A) UMAP embedding coloured and labeled by cell line identity.
 H1975 cells (orange) show two distinct groups reflecting an internal heterogeneity.
-(B) UMAP embedding showing cells colored by level 1 of the hierarchy proposed by the nested Stochastic Block Model.
-(C) UMAP embedding showing cells colored according to the Leiden method at resolution $\gamma = 1$.
-(D) UMAP embedding colored by the classification made by SCCAF when partitions in (B) are used.
-(E) UMAP embedding colored by the classification made by SCCAF when partitions in (C) are used.}\label{FigureTian}
+(B) UMAP embedding showing cells coloured by level 1 of the hierarchy proposed by the nested Stochastic Block Model.
+(C) UMAP embedding showing cells coloured according to the Leiden method at resolution $\gamma = 1$.
+(D) UMAP embedding coloured by the classification made by SCCAF when partitions in (B) are used.
+(E) UMAP embedding coloured by the classification made by SCCAF when partitions in (C) are used.}\label{FigureTian}
 \end{figure}
@@
 
-\subsection*{Hierarchy modeling complies with biological properties}
+\subsection*{Hierarchy modelling complies with biological properties}
 
-When grouping is performed by optimization of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions.
+When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions.
 Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
@@
 Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy.
-Cells are colored according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A).
-In subsequent panels, cells are colored using a signature of Erythroid lineage (B), Monocytes (C) or Neutrophils (D).}\label{Figure_Hemato_Lineage}
+Cells are coloured according to groupings at level 2 of the hierarchy, group 0 marks the most primitive population (A).
+In subsequent panels, cells are coloured using a signature of Erythroid lineage (B), Monocytes (C) or Neutrophils (D).}\label{Figure_Hemato_Lineage}
 \end{figure}
@@
 To do so, we only consider levels with at least two groups, hence excluding the root of the tree.
-We tested this metric on four datasets from \cite{mereu_2020} with different quality levels (iCELL8, MARS-seq, 10XV3 and Quartszeq-2) (Fig.
+We tested this metric on four datasets from \cite{mereu_2020} with different quality levels (iCELL8, MARS-seq, 10XV3 and Quartsz-seq2) (Fig.
 \ref{Figure_Stability}).
@@
 We sought to exploit this property to address the task of annotating cells according to a reference sample.
-To this end we analyzed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies.
+To this end we analysed datasets from \cite{mereu_2020}, which includes mixtures of human PBMC and HEK293T cells profiled with various technologies.
 We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartz-seq2 or MARS-seq.
@@
 \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927.
-When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabeled cells.
+When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabelled cells.
 This misannotation is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transferred, resulting in the lowest accuracy (0.243).
@@
 
-A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalized count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution.
-According to this strategy, the optimal level $i^k$ is the one that minimizes the number of partitions and $\hat{k}$:
+A possible strategy is based on Random Matrix Theory, as suggested by the authors of the SC3 package  \cite{kiselev_2017}, for which a suitable number of clusters, $\hat{k}$, is determined by the number of eigenvalues of the $\boldsymbol Z^\top\boldsymbol Z$ matrix (where $\boldsymbol Z$ is the normalised count matrix) significantly different at $p<.001$ from the appropriate Tracy-Widom distribution.
+According to this strategy, the optimal level $i^k$ is the one that minimises the number of partitions and $\hat{k}$:
 
@@
 An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels.
-While nSBM does not optimize the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is
+While nSBM does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is
 
@@
 
-Where $Q_x$ is modularity at level $x$. We collected values arising from both the approaches for some datasets used in this work (Table \ref{table_optimal})
+Where $Q_x$ is modularity at level $x$. We collected values arising from both the approaches for some datasets used in this work (Table \ref{table_optimal} and Fig. \ref{FigureOptimalLevel})
 
@@
 While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges.
-Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged.
+Moreover, to collect a consensus partition, we minimise multiple models (default: 100) that need to be averaged.
 To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM).
@@
 Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}.
-Times are expressed in mm:ss.
+Times are expressed in MM:SS.
 }
@@
 As observed, model minimisation may be hundred times slower than the extremely fast Leiden approach.
-It should be noted that \texttt{schist} initializes multiple models that are treated by multiple concurrent processes.
-\texttt{graph-tool} itself supports CPU-level parallelization for some of its tasks.
-These optimizations are well suited for clustered computing infrastructure.
-Further development, possibly including GPU-level parallelization, is surely required to accomodate the large size of datasets that are being produced.
+It should be noted that \texttt{schist} initialises multiple models that are treated by multiple concurrent processes.
+\texttt{graph-tool} itself supports CPU-level parallelisation for some of its tasks.
+These optimisations are well suited for clustered computing infrastructure.
+Further development, possibly including GPU-level parallelisation, is surely required to accomodate the large size of datasets that are being produced.
 
@@
 Unless differently stated, all the analysis were produced using \texttt{scanpy} v1.7.1 \cite{wolf_2018} and \texttt{schist} v0.7.6 and the corresponding dependencies.
-All models were initialized 100 times, herein including Leiden partitioning for which we also calculated the consensus partition.
+All models were initialised 100 times, herein including Leiden partitioning for which we also calculated the consensus partition.
 
@@
 Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out.
-\emph{k}NN graph was built with default parameters (50 components and 15 nearest neighbors).
+\emph{k}NN graph was built with default parameters (50 components and 15 nearest neighbours).
 Data were assessed by SCCAF using cell line annotation.
@@
 Data were retrieved using \texttt{squidpy.datasets.visium\_hne\_adata()} built-in function, without further processing.
-Leiden clustering was performed using \texttt{schist.inference.leiden()} function, allowing for 100 initializations, with resolutions $\gamma = 1$ and $\gamma = 0.5$.
+Leiden clustering was performed using \texttt{schist.inference.leiden()} function, allowing for 100 initialisations, with resolutions $\gamma = 1$ and $\gamma = 0.5$.
 
@@
 Genes that were expressed in less than five cells were removed.
-Data were normalized and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5.
+Data were normalised and log-transformed, highly variable genes were detected at minimal dispersion equal to 0.5.
 Neighbourhood graph was built using 30 principal components and 20 neighbours.

### 2021-04-27 supplementary.tex
Final retouch

@@
 \caption[]{Analysis of a mixture of 5 known cell lines.
-(A) UMAP  embedding of cell lines from the sc-mixology experiment colored by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding colored by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
+(A) UMAP  embedding of cell lines from the sc-mixology experiment coloured by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding coloured by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
 \end{figure}
@@
 \caption[]{Analysis of hematopoietic differentiation.
-(A) Low dimensional embedding of single cells colored by original cell type and pseudotime.
-(B) Cells are colored according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
+(A) Low dimensional embedding of single cells coloured by original cell type and pseudotime.
+(B) Cells are coloured according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
 \end{figure}
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Paul15_Leiden_r.png}
-\caption[]{Low dimension embedding of single cells for hematopoietic differentiation colored according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
+\caption[]{Low dimension embedding of single cells for hematopoietic differentiation coloured according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.
 Lowering the distribution does not grant that cells are grouped in a hierarchical way, \emph{e.g.} groups 6 and 8 in the Erythroid branch at resolution $\gamma=1$ are merged or split at coarser resolutions ($\gamma=0.6$ and $\gamma=0.3$).}\label{Figure_Paul15_Leiden_r}
@@
 \includegraphics[keepaspectratio,width=\textwidth,width=0.50\textheight]{Figure_Stability.png}
-\caption[]{Cell Stability. UMAP embeddings of PBMC data profiled with four different technologies ranked by their quality. Cells are colored by the Cell Stability metric.}\label{Figure_Stability}
+\caption[]{Cell Stability. UMAP embeddings of PBMC data profiled with four different technologies ranked by their quality. Cells are coloured by the Cell Stability metric.}\label{Figure_Stability}
 \end{figure}
@@
 \caption[]{Choice of the optimal hierarchy level.
-Datasets analyzed for choice of the most relevant hierarchy level are represented here, from top to bottom in the same order as in Table \ref{table_optimal}.
+Datasets analysed for choice of the most relevant hierarchy level are represented here, from top to bottom in the same order as in Table \ref{table_optimal}.
 For each dataset we report the UMAP embedding coloured by the annotation given in the corresponding manuscript, the profile of modularity $Q$ at different level of the hierarchy and the UMAP embeddings coloured by the level according to $i^k$ or $i^Q$.

### 2021-04-27 manuscript.tex
Last minute mods

@@
 To overcome these issues, \texttt{schist} fits multiple instances in parallel and returns the inferred consensus model, alongside the marginal probabilities for each cell to belong to a specific group (\emph{cell marginals}).
-In addition to the default nested Stochastic Block Model, \texttt{schist} implements the possibility to fit a Bayesian model able to find statistically significant assortative modules \cite{Zhang_Peixoto_2020}, eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.
+Moreover, the Stochastic Block Model has no constraints on what type of modular structure is fitted, meaning that groups are not necessarily identified only by assortativity (\emph{i.e.} cells are mostly connected within the same group).
+When assortativity is thought to be the dominant pattern another model (the Planted Partition Block Model, PPBM \cite{Zhang_Peixoto_2020}), also implemented in \texttt{schist}, is better suited to find statistically significant assortative communities, also eliminating the need to set a resolution parameter as required in standard community detection by maximisation of modularity.
 
-Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
+%Once a model has been fitted, \texttt{schist} allows to evaluate the difference in entropy given by assigning every cell to every possible group.
 This step allows to estimate the strength of association of a cell to a specific group (\emph{cell affinity}).
@@
 In each merged dataset, we retained cell type annotations for 10X cells, while we assigned a "Unknown" label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartz-seq2).
-We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown".
+We then calculated the \emph{cell affinity} matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or "Unknown".
 Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood.

### 2021-05-20 manuscript.tex
BMC amendments

@@
 \title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
+
 \author[1,2]{Leonardo Morelli}
 \author[1,3]{Valentina Giansanti}
-\author[1]{Davide Cittaro}
+\author[1,\thanks{Corresponding author: cittaro.davide@hsr.it}]{Davide Cittaro}
+
 \affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
@@
 To show this property we performed a simple experiment on a randomised \emph{k}NN graph.
-We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged.
+We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}, Fig.
+\ref{FigureRandom}A) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged.
 We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$).
-We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.
+We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group (Fig.
+\ref{FigureRandom}B).
 
@@
 
-\section*{Acknowledgements}
+\section*{Declarations}
 
+\subsection*{Ethics approval and consent to participate}
+Not applicable
+
+\subsection*{Consent for publication}
+Not applicable
+
+\subsection*{Availability of data and materials}
+No datasets were generated in the current study.
+The original third-party datasets that were analysed are included in the corresponding publications \cite{plass_2018,  mereu_2020, Tian_2019, paul_2015, Zeisel_2015, Bastidas_Ponce_2019, Baron_2016, Aizarani_2019}
+
+\subsection*{Competing interests}
+The authors declare that they have no competing interests.
+
+\subsection*{Funding}
+This work has been supported by Accelerator Award: A26815 entitled: ”Single-cell cancer evolution in the clinic” funded through a partnership between Cancer Research UK and Fondazione AIRC.
+
+\subsection*{Authors' contributions}
+L.M., V.G. and D.C. performed the analysis and wrote the manuscript. L.M. and D.C wrote the software. D.C. conceived the study.
+
+\subsection*{Acknowledgements}
 We would like to thank Tiago de Paula Peixoto (Central European University, ISI Foundation) and Giovanni Petri (ISI Foundation) for the discussions and the precious hints.
 We also would like to thank all people at COSR, in particular Giovanni Tonon and Paolo Provero.
-This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC.
 

### 2021-07-08 supplementary.tex
modified fig s6 size
Fixed typo in legend of Fig S2

@@
 (A) Low dimensional embedding of single cells coloured by original cell type and pseudotime.
-(B) Cells are coloured according to the nSBM grouping at level 3 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
+(B) Cells are coloured according to the nSBM grouping at level 1 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
 \end{figure}
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=1\textheight]{FigureOptimalLevel.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.9\textheight]{FigureOptimalLevel.png}
 \caption[]{Choice of the optimal hierarchy level.

### 2021-07-11 manuscript.tex
addressing reviewer 1, comment 3
addressing reviewer 1, comment 4

@@
 
-When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, that is communities are consistently grouped at lower resolutions.
+When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \emph{i.e.} communities are consistently grouped at lower resolutions.
 Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
@@
 The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself.
-Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data.
+Differences in entropy can be used to perform model selection, hence we can choose which model better describes the data.
 We sought to exploit this property to address the task of annotating cells according to a reference sample.
@@
 \centering
- \begin{tabular}{|| l c c c c c ||}
+ \begin{tabular}{|| l c c | c c c ||}
 %  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}

### 2021-07-12 manuscript.tex
Red text for previous changes

@@
 
-When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \emph{i.e.} communities are consistently grouped at lower resolutions.
+When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \textcolor{red}{\emph{i.e.}} communities are consistently grouped at lower resolutions.
 Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
@@
 The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme, not limited to the solution given by the model itself.
-Differences in entropy can be used to perform model selection, hence we can choose which model better describes the data.
+Differences in entropy can be used to perform model selection, \textcolor{red}{hence} we can choose which model better describes the data.
 We sought to exploit this property to address the task of annotating cells according to a reference sample.

### 2021-07-12 supplementary.tex
Red text for previous changes

@@
 (A) Low dimensional embedding of single cells coloured by original cell type and pseudotime.
-(B) Cells are coloured according to the nSBM grouping at level 1 of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
+(B) Cells are coloured according to the nSBM grouping at level \textcolor{red}{1} of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
 \end{figure}

### 2021-07-15 manuscript.tex
updated references
Initial work on random data

@@
 For relatively small datasets (\emph{i.e.} less than 10k cells), the first level of the hierarchy contains a number of groups in line with how many observable populations are.
+\textcolor{red}{Notwithstanding, cell groups identified at leach level may have a biological interpretation.
+In particular, groups at deepest levels (0 or 1) may be relevant when studying rare populations.
+For example, in the hematopoiesis dataset shown in Fig.
+\ref{Figure_Hemato_Supp}A, groups 11 (DC) and 19 (Lymph) cannot be distinguished from nSBM level 1 and up; a closer investigation to level 0, however, revealed that these cells are clearly separated.
+To better understand the role of deepest levels, we performed an additional analysis of a single cell dataset of mouse crypt cells \cite{Ayyaz_Kumar_2019}, which was also covered in a recent paper proposing GapClust as an optimal approach to identify rare cell populations \cite{Fa_Wei_Zhou_Johnston_2021}.
+We sought to identify the four rare populations identified by GapClust.
+We could identify all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig.
+FOOBAR), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
+Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig.SUPPFOOBAR), not even at high resolution, hence prompting the development of specific approaches such as GapClust.
+As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we also investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ), again using the Crypt experiment as representative sample.
+We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction.
+Looking at the agreement among different partitions, we found that it is consistent among different solutions, both at level 0 and level 1, with the exception of models for graph derived with low number of PC (Fig.
+ABRACADABRA)}
 %\subsection*{PPBM?}

### 2021-07-15 manuscript.tex
Initial work on random data

@@
 
-\subsection*{Absence of information is correctly handled by nSBM}
+\subsection*{\texttt{schist} is robust to data perturbation}
 
@@
 Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
+\textcolor{red}{To understand the impact of random noise on structured data, we considered the same dataset and added gaussian noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of expressed genes (\emph{e.g.} genes with counts higher than zero).
+We noticed that \texttt{schist} is more robust to perturbations, although none of the methods can be considered reliable when $\sigma \geq 1.5$ (Table \ref{Table_Noise}.}
 

### 2021-07-15 manuscript.tex
Removed TIAN0

--- 2021-07-15 manuscript.tex
@@
 
-\subsection*{Absence of information is correctly handled by nSBM}
+\subsection*{\texttt{schist} is robust to data perturbation}
 
-One of the most relevant difference between SBM and other methods to cluster single cells is that it relies on robust statistical modelling.
+One of the most relevant difference between the SBM and other methods to cluster single cells is that it relies on robust statistical modelling.
 In this sense, the number of groups identified strictly mirrors the amount of information contained in the data.
-An important consequence is that absence of information can be properly handled.
+An important consequence is that absence of information \textcolor{red}{(\emph{i.e.} maximal entropy)} can be properly handled.
 To show this property we performed a simple experiment on a randomised \emph{k}NN graph.
@@
 We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$).
-We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group (Fig.
+We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group \textcolor{red}{at default resolution} (Fig.
 \ref{FigureRandom}B).
@@
 
-This experiment is a deliberate extreme case.
+\textcolor{red}{Only by reducing resolution to $\gamma \leq 0.5$ we were able to obtain a single partition by modularity (Fig.
+\ref{Modularity_Random_Res}).
+Of course,} this experiment is a deliberate extreme case.
 The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information.
@@
 Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
+\textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes.
+We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}.
+Of note, the concordance with original annotations drops at $\sigma \geq 1.5$ for both the approaches.}
 
@@
 \ref{FigureTian}E), highlighting potential limitations of this approach.
-In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig.
+%In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig.
 \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig.

### 2021-07-15 supplementary.tex
Initial work on random data
Removed TIAN0

@@
 \renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 
@@
 \maketitle
+
+\section*{Supplementary tables}
+\begin{table}[h!]
+\centering
+ \begin{tabular}{||c|c c|c c||}
+%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
+ \hline
+\multicolumn{1}{||c|}{\textbf{Noise level ($\sigma$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\
+
+ & $ARI$ & $N$ & $ARI$ & $N$  \\ 
+ \hline\hline
+-- & 0.9228 & 8 & 0.8944 & 6 \\
+\hline 
+0.5 & 0.6249 & 10 & 0.7509  & 10 \\ 
+ \hline
+ 1.0  & 0.5966 & 7 & 0.7269  & 6 \\
+ \hline
+ 1.5  & 0.2833 & 9 &0.3029 & 3 \\
+ \hline
+ 2.0  & 0.1405 & 12 & 0.1961 & 2 \\
+ \hline
+\end{tabular}
+\caption{\textcolor{red}{Effect of random noise }}
+\label{Table_Noise}
+\end{table}
+
+\clearpage
+
 \section*{Supplementary figures}
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Tian_NSBM0.png}
-\caption[]{Analysis of a mixture of 5 known cell lines.
-(A) UMAP  embedding of cell lines from the sc-mixology experiment coloured by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding coloured by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.375\textheight]{Random_Leiden_by_gamma.png}
+\caption[]{\textcolor{red}{foobarbaz}}
+\label{Modularity_Random_Res}
 \end{figure}
 \clearpage
+
+%\begin{figure}[htbp]
+%\centering
+%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Tian_NSBM0.png}
+%\caption[]{Analysis of a mixture of 5 known cell lines.
+(A) UMAP  embedding of cell lines from the sc-mixology experiment coloured by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding coloured by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
+%\end{figure}
+%\clearpage
 %\begin{figure}[htbp]

### 2021-07-15 manuscript.tex
Random data in discussion

@@
 We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques.
-The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution.
+\textcolor{red}{As  random data may have modular structure \cite{guimer_2004}, an important property of our approach is that it does not overfit data by finding partitions when, in fact, the are not.} Another important advantage is \textcolor{that} the hierarchical definition of cell groups eliminates the choice of an arbitrary threshold on clustering resolution.
 In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.

### 2021-07-15 manuscript.tex
Modified Figure 1

--- 2021-07-15 manuscript.tex
@@
 \includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.png}
-\caption[]{nSBM results for randomised data.
+\caption[]{Modeling randomised data.
 (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right).
@@
 \centering
-\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.png}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.pdf}
 \caption[]{Analysis of a mixture of 5 known cell lines.
@@
 We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques.
-The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution.
+\textcolor{red}{As  random data may have modular structure \cite{guimer_2004}, an important property of our approach is that it does not overfit data by finding partitions when, in fact, the are not.} Another important advantage is \textcolor{red}{that} the hierarchical definition of cell groups eliminates the choice of an arbitrary threshold on clustering resolution.
 In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.

### 2021-07-16 manuscript.tex
use eps from now on
Fig 1 eps
I make my own pictures now,from source...
Another PDF figure

@@
 \centering
-\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.png}
+\includegraphics[keepaspectratio,width=0.5\textwidth,height=\textheight]{FIgure_Random.pdf}
 \caption[]{Modeling randomised data.
@@
 \centering
-\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FigureTian.pdf}
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{FigureTian.pdf}
 \caption[]{Analysis of a mixture of 5 known cell lines.
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_Visium_Spatial.png}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Visium_Spatial.pdf}
 \caption[]{Analysis of spatial transcriptomics of a coronal section of mouse brain by Visium H\&E.
@@
 \centering
-\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Hemato_Lineage.png}
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Hemato_Lineage.pdf}
 \caption[]{Analysis of hematopoietic differentiation.

### 2021-07-16 supplementary.tex
Another PDF figure
added other

@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.pdf}
 \caption[]{Analysis of hematopoietic differentiation.
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Paul15_Leiden_r.png}
+\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Paul15_Leiden_r.pdf}
 \caption[]{Low dimension embedding of single cells for hematopoietic differentiation coloured according to Leiden clustering at decreasing resolution, from 1.0 to 0.1.

### 2021-07-16 manuscript.tex
Another one
last complex figure
added figure for tuft cells

@@
 \includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Visium_Spatial.pdf}
+\vspace*{-80pt}
 \caption[]{Analysis of spatial transcriptomics of a coronal section of mouse brain by Visium H\&E.
@@
 \centering
-\includegraphics[keepaspectratio,width=0.9\textwidth,height=\textheight]{Figure_Label_Transfer.png}
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Label_Transfer.pdf}
+\vspace*{-40pt}
 \caption[]{Label transfer using SBM.
@@
 We could identify all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig.
-FOOBAR), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
-Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig.SUPPFOOBAR), not even at high resolution, hence prompting the development of specific approaches such as GapClust.
-As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we also investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ), again using the Crypt experiment as representative sample.
+\ref{Figure_Rare_Main}), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
+Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig.
+\ref{Figure_Rare_Supp}), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}
+
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=0.5\textwidth,height=\textheight]{Figure_Rare_Cells_Main.pdf}
+%\vspace*{-40pt}
+\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Main}
+\end{figure}
+
+\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we also investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ), again using the Crypt experiment as representative sample.
 We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction.

### 2021-07-16 manuscript.tex
adjustments

@@
 \centering
-\includegraphics[keepaspectratio,width=0.5\textwidth,height=\textheight]{FIgure_Random.pdf}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.pdf}
 \caption[]{Modeling randomised data.
@@
 \begin{figure}[H]
-\centering
-\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{FigureTian.pdf}
+%\centering
+\raggedleft
+\includegraphics[keepaspectratio,width=1\textwidth]{FigureTian.pdf}
 \caption[]{Analysis of a mixture of 5 known cell lines.
@@
 
-\begin{table}[h!]
+\begin{table}[H]
 \centering
@@
 \centering
-\includegraphics[keepaspectratio,width=0.5\textwidth,height=\textheight]{Figure_Rare_Cells_Main.pdf}
+\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Rare_Cells_Main.pdf}
 %\vspace*{-40pt}

### 2021-07-18 manuscript.tex
added last figure

@@
 
-\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we also investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ), again using the Crypt experiment as representative sample.
+\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ).
 We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction.
@@
 The major drawback of adopting this strategy is the substantial increase of runtimes.
-As observed, model minimisation may be hundred times slower than the extremely fast Leiden approach.
+As observed, model minimisation is many times slower than the extremely fast Leiden approach.
 It should be noted that \texttt{schist} initialises multiple models that are treated by multiple concurrent processes.
@@
 
+\subsection*{Analysis of crypt cells data}
+\textcolor{red}{Count matrix for untreated crypt cells (GSM3308718) was downloaded from GEO.
+Cells with less than 200 genes and genes detected in less than 2 cells were excluded from the analysis.
+After normalization and log-transformation, highly variable genes were selected with a cutoff on the mean expression equal to 0.05.
+Rare subpopulations were first highlighted using the gene signatures published in \cite{Fa_Wei_Zhou_Johnston_2021}:
+}
+\begin{itemize}
+\item R1\_1: Cd8a, Cd3g, Ccl5, Gzma, Gzmb, RGs1, Nkg7, Cd7, Fcer1g
+\item R1\_2: H2-Aa, H2-Ab1, H2-Eb1, Cd74, Ly6d, Ebf1, Cd79a, Mef2c
+\item R2: Krt18, Cd24a, Adh1, Cystm1, Aldh2, Dclk1, Sh2d6, Rgs13, Hck, Trpm5
+\item R3: Alas2, Hbb-bs, Hba-a1, Hbb-bt
+\end{itemize}
+
 \subsection*{Label transfer}

### 2021-07-18 supplementary.tex
last complex figure
added figure for tuft cells
added last figure

@@
 \caption[]{Distribution of second choices in label transfer.
-For each dataset in which we assessed accuracy of label transfer by SMB, we collect the rank of affinities of the correct class when an assignment could not be performed (\emph{i.e.} the highest affinity was for the "Unknonwn" label)  }\label{hist_label_second}
+For each dataset in which we assessed accuracy of label transfer by SBM, we collect the rank of affinities of the correct class when an assignment could not be performed (\emph{i.e.} the highest affinity was for the "Unknonwn" label)  }\label{hist_label_second}
 \end{figure}
@@
 \centering
-\includegraphics[keepaspectratio,width=\textwidth,height=0.9\textheight]{FigureOptimalLevel.png}
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{FigureOptimalLevel.pdf}
+\vspace*{-80pt}
 \caption[]{Choice of the optimal hierarchy level.
@@
 
+\clearpage
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Rare_Cells_Supp.pdf}
+%\vspace*{-40pt}
+\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Supp}
+\end{figure}
+
+\clearpage
+\begin{figure}[H]
+\centering
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel0.pdf} 
+        \caption{Chromium 10x} \label{cm_10x}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel1.pdf} 
+        \caption{iCell8} \label{cm_icell8}
+    \end{subfigure}
+    \hfill
+%    \vspace{40pt}
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel2.pdf} 
+        \caption{Quartz-seq2} \label{cm_quartz}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel3.pdf} 
+        \caption{MARS-seq} \label{cm_mars}
+    \end{subfigure}
+%    \vspace{40pt}
+   \hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel4.pdf} 
+        \caption{Crypt cells} \label{cm_crypt}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{ClusterMap_L0_Panel5.pdf} 
+        \caption{Planaria} \label{cm_planaria}
+    \end{subfigure}
+    \vspace{40pt}
+
+%\vspace*{-40pt}
+\caption[]{Consistency at Level 0} \label{Figure_Clustermap_L0}
+\end{figure}
+
 \end{document}

### 2021-07-18 manuscript.tex
Final ones

@@
 We could identify all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig.
-\ref{Figure_Rare_Main}), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
+\ref{Figure_Rare_Main} and \ref{Figure_Rare_Supp}), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
 Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig.
-\ref{Figure_Rare_Supp}), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}
+\ref{Figure_Rare_Supp}C), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}
 
@@
 
-\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig SUPPBAZBAZ).
+\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig.
+\ref{Figure_Comm_Count} and \ref{Figure_Comm_Size}).
 We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction.
 Looking at the agreement among different partitions, we found that it is consistent among different solutions, both at level 0 and level 1, with the exception of models for graph derived with low number of PC (Fig.
-ABRACADABRA)}
+\ref{Figure_Clustermap_L0})}
 %\subsection*{PPBM?}

### 2021-07-18 supplementary.tex
Final ones

@@
 \centering
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Community_Count_L0.pdf}
+%\vspace*{-40pt}
+\caption[]{Number of communities} \label{Figure_Comm_Count}
+\end{figure}
+
+\clearpage
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Community_Size_L0.pdf}
+%\vspace*{-40pt}
+\caption[]{Size of communities} \label{Figure_Comm_Size}
+\end{figure}
+
+\clearpage
+\begin{figure}[H]
+\centering
     \begin{subfigure}[t]{0.3\textwidth}

### 2021-07-19 manuscript.tex
Update manuscript.tex

@@
 Nevertheless, real-world data may include an unknown amount of random noise.
-Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
+Hence, it is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
 \textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes.

### 2021-07-19 manuscript.tex
Update manuscript.tex

@@
 \textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes.
-We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}.
+We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}).
 Of note, the concordance with original annotations drops at $\sigma \geq 1.5$ for both the approaches.}

### 2021-07-19 manuscript.tex
Update manuscript.tex

@@
 Looking at the agreement among different partitions, we found that it is consistent among different solutions, both at level 0 and level 1, with the exception of models for graph derived with low number of PC (Fig.
-\ref{Figure_Clustermap_L0})}
+\ref{Figure_Clustermap_L0}).}
 %\subsection*{PPBM?}
@@
 We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques.
-\textcolor{red}{As  random data may have modular structure \cite{guimer_2004}, an important property of our approach is that it does not overfit data by finding partitions when, in fact, the are not.} Another important advantage is \textcolor{red}{that} the hierarchical definition of cell groups eliminates the choice of an arbitrary threshold on clustering resolution.
+\textcolor{red}{As random data may have modular structure \cite{guimer_2004}, an important property of our approach is that it does not overfit data by finding partitions when, in fact, there are not.} Another important advantage is \textcolor{red}{that} the hierarchical definition of cell groups eliminates the choice of an arbitrary threshold on clustering resolution.
 In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.

### 2021-07-20 manuscript.tex
ongoing fixes

@@
 We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$).
-We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group \textcolor{red}{at default resolution} (Fig.
+We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups \textcolor{red}{at default resolution}, whereas \textcolor{red}{\texttt{schist}} does not identify any cell group, \textcolor{red}{at level 0} (Fig.
 \ref{FigureRandom}B).
@@
 (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges.
-In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
+In this case \texttt{schist} does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
 \end{figure}
 
-\textcolor{red}{Only by reducing resolution to $\gamma \leq 0.5$ we were able to obtain a single partition by modularity (Fig.
+\textcolor{red}{Only by reducing resolution to $\gamma < 0.6$ we were able to obtain a single partition by modularity (Fig.
 \ref{Modularity_Random_Res}).
@@
 Hence, it is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.
-\textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes.
+\textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modelled after the feature-wise distribution of detected genes.
 We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}).
@@
 \ref{FigureTian}C).
-The nSBM solution correctly identifies H1975 groups as a single entity at level 1 of the hierarchy.
+\texttt{schist} correctly identifies H1975 groups as a single entity at level 1 of the nSBM hierarchy.
 We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes.
@@
 When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \textcolor{red}{\emph{i.e.}} communities are consistently grouped at lower resolutions.
-Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree.
+Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas a nSBM inherently encodes hierarchies by merging communities in a tree.
 The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.
@@
 
-\subsection*{Analysis of Random data}
+\subsection*{Analysis of Randomized data}
 
@@
 UMAP embedding was recomputed after randomisation using the shuffled graph.
+\textcolor{red}{To generate data with white noise we computed the genewise means ($\mu_g$) and standard deviations ($\sigma_g$) of log-normalized counts excluding 0 values.
+We generated random values using $\mu_g$ and $k\sigma_g$, $k \in \{0.5, 1, 1.5, 2\}$, and added to original expression values.
+}
 

### 2021-07-20 manuscript.tex
final changes

@@
 \ref{Figure_Visium}).
-When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.
+When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas \texttt{schist} correctly identifies the pyramidal layer.
 
@@
 Tissues are well defined at default resolution for the standard approach.
-When resolution is decreased to $\gamma = 0.5$, cells from the dentate gyrus are merged to the hippocampus.
-When nSBM is used, the structure of the pyramidal is maintained at different levels of the hierarchy}\label{Figure_Visium}
+When resolution is decreased to $\gamma = 0.5$, cells are aggregated breaking the histological types, \emph{e.g.} cells from the dentate gyrus are merged to the hippocampus.
+When a nSBM is applied, the structure of the pyramidal layer is maintained at different levels of the hierarchy}\label{Figure_Visium}
 \end{figure}
@@
 \ref{Figure_Hemato_Lineage}).
-Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime.
+Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime (Fig.
+\ref{Figure_Hemato_Supp}B).
 Conversely, the Leiden method at default resolution identified 24 groups.
@@
 The first line reports UMAP embeddings for datasets profiled with Chromium 10X V3, Quartz-seq2 and  MARS-seq, each annotated by known cell types.
-Quartz-seq2 and MARS-seq were reannotated using \emph{k}NN method, \texttt{sc.tl.ingest() } or \texttt{schist}.
+Quartz-seq2 and MARS-seq were reannotated using \emph{k}NN method, \texttt{sc.tl.ingest()} or \texttt{schist}.
 The accuracy of each label transfer task is reported above the corresponding UMAP.} \label{Figure_Label_Transfer}
@@
 An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels.
-While nSBM does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is
+While \texttt{schist} does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is
 
@@
 For example, in the hematopoiesis dataset shown in Fig.
-\ref{Figure_Hemato_Supp}A, groups 11 (DC) and 19 (Lymph) cannot be distinguished from nSBM level 1 and up; a closer investigation to level 0, however, revealed that these cells are clearly separated.
+\ref{Figure_Hemato_Supp}A, groups 11 (DC) and 19 (Lymph) cannot be distinguished from nSBM level 1 and up; a closer investigation to level 0, however, revealed that these cells are clearly separated (Fig.
+\ref{Figure_Hemato_Supp}D).
 To better understand the role of deepest levels, we performed an additional analysis of a single cell dataset of mouse crypt cells \cite{Ayyaz_Kumar_2019}, which was also covered in a recent paper proposing GapClust as an optimal approach to identify rare cell populations \cite{Fa_Wei_Zhou_Johnston_2021}.
 We sought to identify the four rare populations identified by GapClust.
-We could identify all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig.
-\ref{Figure_Rare_Main} and \ref{Figure_Rare_Supp}), suggesting that studying levels with appropriate community size is a valid method to spot rare populations.
-Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig.
+We could distinguish all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig.
+\ref{Figure_Rare_Main} and \ref{Figure_Rare_Supp}), suggesting that exploring nSBM levels with appropriate community size is a valid method to spot rare populations.
+Of note, modularity optimisation could not pinpoint Tuft cells in appropriate way (Fig.
 \ref{Figure_Rare_Supp}C), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}
@@
 %\vspace*{-40pt}
-\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Main}
+\caption[]{Identification of rare populations.
+UMAP embeddings of mouse crypt cells highlighting the groups found at level 1 of nSBM hierarchy corresponding to rare populations identified by GapClust \cite{Fa_Wei_Zhou_Johnston_2021}.
+Each rare population is colored by a signature calculated from published gene lists (see methods).} \label{Figure_Rare_Main}
 \end{figure}
@@
 \ref{Figure_Comm_Count} and \ref{Figure_Comm_Size}).
-We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction.
-Looking at the agreement among different partitions, we found that it is consistent among different solutions, both at level 0 and level 1, with the exception of models for graph derived with low number of PC (Fig.
-\ref{Figure_Clustermap_L0}).}
+We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, with different solutions being consistent (Fig.
+\ref{Figure_Clustermap_L0}), suggesting that higher number of neighbors provides richer description of the dataset.
+The number of PCs used to evaluate cell-to-cell distance influences the variability of community sizes; the consistency among different solutions is high when a sufficient number of PCs is chosen, data suggest that for large datasets more PCs should be included to include adequate fraction of overall variability.}
 %\subsection*{PPBM?}
@@
 
-Minimisation of the nSBM is a process that requires a large amount of computational resources.
+Minimisation of a nSBM is a process that requires a large amount of computational resources.
 While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges.

### 2021-07-20 supplementary.tex
final changes

--- 2021-07-18 supplementary.tex
@@
  \hline
-\multicolumn{1}{||c|}{\textbf{Noise level ($\sigma$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\
+\multicolumn{1}{||c|}{\textbf{Noise level ($k$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\
 
@@
 \end{tabular}
-\caption{\textcolor{red}{Effect of random noise }}
+\caption{Effect of random noise.
+The table shows performance of Leiden approach and nested Stochastic Block Model in recognising annotated population in PBMC 3k data when increasing level of noise are added to log-normalized counts.
+The $k$ parameter multiplies the $\sigma$, estimated from data, to generate random counts.
+$ARI$: Adjusted Rand Index, $N$: number of communities.}
 \label{Table_Noise}
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.375\textheight]{Random_Leiden_by_gamma.png}
-\caption[]{\textcolor{red}{foobarbaz}}
+\caption[]{Decreasing the resolution parameter $\gamma$ results in lower number of communities also in the random graph generated from PBMC 3k data.
+At $\gamma < 0.5$, Leiden approach correctly identifies one single community.}
 \label{Modularity_Random_Res}
@@
 \includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.pdf}
-\caption[]{Analysis of hematopoietic differentiation.
-(A) Low dimensional embedding of single cells coloured by original cell type and pseudotime.
-(B) Cells are coloured according to the nSBM grouping at level \textcolor{red}{1} of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
+\caption[]{Low dimensional embedding of single cells from hematopoietic differentiation dataset.
+Cells are colored by original cell type (A), pseudotime (B), nSBM grouping at level \textcolor{red}{1} of the hierarchy (C) and level 0 (D).
+Since level 0 identifies 114 groups, only group labels have been represented.}\label{Figure_Hemato_Supp}
 \end{figure}
@@
 %\vspace*{-40pt}
-\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Supp}
+\caption[]{Identification of rare populations.
+(A) UMAP embedding of mouse crypt cells coloured by cell groups resulting at level 1 of the nSBM hierarchy.
+(B) Cells coloured by the signature of R3 group identified by GapClust.
+(C) UMAP embedding of mouse crypt cells coloured after optimisation of modularity at default parameters (left) or higher resolution (right).} \label{Figure_Rare_Supp}
+\end{figure}
+
+\clearpage
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=.95\textwidth,height=\textheight]{Community_Count_L0.pdf}
+%\vspace*{-40pt}
+\caption[]{Evaluation of the number of communities.
+Barplots show the number of communities found at level 0 of the nSBM hierarchy for various datasets used in the manuscript when different parameters are chosen to build the \emph{k}NN graph.
+Increasing the number of neighbors results in more granular partitioning (\emph{i.e.} more communities), whereas increasing the number of principal components used to evaluate cell-to-cell distance generally works in opposite way.
+$x$-axis: number of principal components; $y$-axis: number of communities; bar color: number of \emph{k}NN neighbors.} \label{Figure_Comm_Count}
+\end{figure}
+
+\clearpage
+\begin{figure}[H]
+\centering
+\includegraphics[keepaspectratio,width=.95\textwidth,height=\textheight]{Community_Size_L0.pdf}
+%\vspace*{-40pt}
+\caption[]{Evaluation of the size of communities.
+Boxplots show the distribution of size of communities found at level 0 of the nSBM hierarchy for various datasets used in the manuscript when different parameters are chosen to build the \emph{k}NN graph.
+Increasing the number of neighbors results in more granular partitioning (\emph{i.e.} smaller communities), increasing the number of principal components results in more dispersed sizes.
+$x$-axis: number of principal components; $y$-axis: community size; bar color: number of \emph{k}NN neighbors.} \label{Figure_Comm_Size}
 \end{figure}
@@
 %\vspace*{-40pt}
-\caption[]{Consistency at Level 0} \label{Figure_Clustermap_L0}
+\caption[]{Consistency at Level 0.
+Each heatmap shows the Adjusted Rand Index among partitions found when different parameters are used to build th \emph{k}NN graph.
+Row labels encode the number of PCs and the number of neighbors (P\_N), reflected in colors used to annotate columns.} \label{Figure_Clustermap_L0}
 \end{figure}
